---
title: "Direct and Indirect Effects of Coda Voicing on Onset VOT: A Bayesian Mediation Analysis"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Overview

This analysis tests whether coda voicing exerts an *indirect* effect
on onset voice onset time (VOT), mediated by vowel duration:

$$\text{Coda Voicing} \xrightarrow{a} \text{Vowel Duration}
\xrightarrow{b} \text{VOT}$$

This document supersedes an earlier draft that used frequentist SEM
(`lavaan`) and a random-intercept-only mediation approach
(`mediation` + `lme4`). It is rewritten here to be consistent with the
primary VOT analysis in every respect that matters for comparability:

- **Units**: `vot` and `vdur` are in milliseconds throughout (the
  earlier draft used seconds).
- **Random-effects structure**: mediation is estimated from two
  separate **Bayesian maximal-structure models** (mediator equation
  and outcome equation), each fit with `brms` using by-subject random
  slopes for place of articulation, coda voicing, and their
  interaction — the same maximal logic used for the primary VOT
  model, applied consistently to both mediation equations. A fully
  joint multivariate version, which links the two equations' random
  effects into one shared correlation matrix, was attempted first but
  proved computationally impractical (78 correlation terms from 20
  subjects); that attempt and the reasoning for stepping back from it
  are documented in Section 5.
- **Priors**: informed priors are used only where independent
  literature genuinely supports a specific magnitude, following the
  same tiered logic used for the primary VOT model's priors (strong
  literature support vs. directional-but-uncertain support vs. no
  informative prior at all for the focal, untested quantity).
- **Inference**: indirect, direct, and total effects are computed by
  multiplying posterior draws from the two separately-fit models. This
  is an approximation — it treats the two equations' random effects as
  independent across equations, when in reality a given speaker's
  mediator-equation and outcome-equation random effects are likely
  correlated. This is a smaller and more defensible simplification
  than the random-intercept-only assumption of the original
  `mediation`/`lme4` approach, and is standard practice when a fully
  joint model is not tractable.

A frequentist comparison (the original `lavaan`/`mediation` approach)
is retained at the end as an optional robustness check, not as the
primary analysis.

---

# 1. Load Packages

```{r libraries}
library(brms)
library(tidyverse)
library(posterior)
```

---

# 2. Prepare the Data

Data preparation matches the primary VOT analysis exactly: `ons_poa`
coded once as a `p`/`t`/`k` factor (reference level `/p/`), `coda_vc`
as a 0/1 factor, and `vot`/`vdur` converted from seconds to
milliseconds.

```{r data-prep}
data <- read.csv("/Users/chandan/Desktop/GitHub/Burst_voice/Data/data_12-11-24.csv",
                  header = TRUE)

data <- data %>%
  mutate(
    ons_poa = factor(ons_poa, levels = c(1, 2, 3), labels = c("p", "t", "k")),
    coda_vc = factor(coda_vc, levels = c(0, 1)),
    vot  = vot  * 1000,
    vdur = vdur * 1000
  )

data_mod <- data %>% filter(!is.na(vot), !is.na(vdur))

number_of_subs <- length(unique(data_mod$sub))
number_of_subs
```

---

# 3. Descriptive Statistics

```{r descriptives}
data_mod %>%
  group_by(ons_poa, coda_vc) %>%
  summarise(
    n         = n(),
    mean_vot  = round(mean(vot),  1),
    mean_vdur = round(mean(vdur), 1),
    .groups   = "drop"
  )
```

```{r plot-descriptives, fig.width=10, fig.height=5}
p1 <- ggplot(data_mod, aes(x = coda_vc, y = vdur, fill = coda_vc)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  facet_wrap(~ons_poa) +
  labs(title = "Vowel Duration by Coda Voicing and Onset Place",
       x = "Coda Voicing", y = "Vowel Duration (ms)") +
  scale_x_discrete(labels = c("Voiceless", "Voiced")) +
  theme_minimal() + theme(legend.position = "none")

p2 <- ggplot(data_mod, aes(x = coda_vc, y = vot, fill = coda_vc)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  facet_wrap(~ons_poa) +
  labs(title = "VOT by Coda Voicing and Onset Place",
       x = "Coda Voicing", y = "VOT (ms)") +
  scale_x_discrete(labels = c("Voiceless", "Voiced")) +
  theme_minimal() + theme(legend.position = "none")

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

---

# 4. Bayesian Mediation Model (primary analysis)

## 4.1 Model structure

The mediator equation (`vdur ~ ons_poa * coda_vc`) and outcome
equation (`vot ~ vdur + ons_poa * coda_vc`) are fit as two **separate**
maximal-structure `brms` models, each with by-subject random slopes
for place of articulation, coda voicing, and their interaction (plus,
in the outcome equation, a random slope for vowel duration itself).
This mirrors exactly the random-effects structure of the primary VOT
model, applied to each mediation path individually.

```{r mediation-formulas}
bf_vdur <- bf(vdur ~ ons_poa * coda_vc + (1 + ons_poa * coda_vc | sub))
bf_vot  <- bf(vot  ~ vdur + ons_poa * coda_vc + (1 + vdur + ons_poa * coda_vc | sub))
```

**Why not fit these jointly?** A fully joint multivariate model
(linking both equations' random effects into one shared correlation
matrix via `brms`' `|q|` syntax) would be the more complete approach,
since it would capture any correlation between a speaker's
mediator-equation and outcome-equation random effects. We attempted
this first; combining both equations' random effects yields 13
correlated by-subject terms (78 unique correlations) from only 20
subjects, and in practice this model did not sample in a practical
amount of time even with an increased `adapt_delta` and
`max_treedepth`. Fitting the two equations separately is a
well-established, more tractable approximation, at the cost of
treating the two equations' random effects as independent across
equations. See Section 5 for the joint model specification and a
discussion of when it might be worth revisiting.

## 4.2 Priors

Priors are tiered by epistemic status, extending the same logic used
for the primary VOT model's priors:

1. **Strong literature support, specific expected magnitude**:
   - Place-of-articulation contrasts on VOT (as in the primary model):
     $\mathcal{N}(15, 8)$ ms for /t/, $\mathcal{N}(30, 12)$ ms for /k/.
   - Coda voicing's effect on vowel duration (path *a*) — this is the
     classic, extensively replicated "voicing effect"
     \citep{Chen1970,House1953,PetersonLehiste1960}. Centered at a
     literature-typical magnitude for English (~55 ms) with a wide SD
     reflecting real cross-study heterogeneity: $\mathcal{N}(55, 30)$ ms.
2. **Directionally established, magnitude uncertain**:
   - Vowel duration's effect on onset VOT (path *b*) — a positive
     relationship is well established directionally, but not tightly
     quantified across studies, so this prior is directional but wide:
     $\mathcal{N}(0.15, 0.15)$.
3. **Focal, untested quantities — neutral, zero-centered**:
   - The *direct* effect of coda voicing on onset VOT, controlling for
     vowel duration (path *c*) — this is the whole point of the
     mediation analysis, so it must not be prejudged:
     $\mathcal{N}(0, 20)$ ms.
   - All place x coda-voicing interaction terms, in both equations.
   - Both intercepts: no informative absolute location (matching the
     primary model's reasoning that absolute baseline durations vary
     too much across speech styles to anchor confidently), constrained
     only to be non-negative via `lb = 0`.

Because the two equations are now fit as separate (single-response)
models, priors no longer need `resp =` qualifiers.

```{r mediation-priors}
# Check before fitting, as always
get_prior(bf_vdur, data = data_mod)
get_prior(bf_vot,  data = data_mod)

vdur_priors <- c(
  prior(normal(0, 150), class = Intercept, lb = 0),
  prior(normal(55, 30), class = b, coef = coda_vc1),
  prior(normal(0, 25),  class = b, coef = "ons_poat:coda_vc1"),
  prior(normal(0, 25),  class = b, coef = "ons_poak:coda_vc1"),
  prior(normal(0, 30),  class = sd),
  prior(normal(0, 30),  class = sigma),
  prior(lkj(2), class = cor)
)

vot_priors <- c(
  prior(normal(0, 100),    class = Intercept, lb = 0),
  prior(normal(15, 8),     class = b, coef = ons_poat),
  prior(normal(30, 12),    class = b, coef = ons_poak),
  prior(normal(0.15, 0.15), class = b, coef = vdur),
  prior(normal(0, 20),     class = b, coef = coda_vc1),
  prior(normal(0, 15),     class = b, coef = "ons_poat:coda_vc1"),
  prior(normal(0, 15),     class = b, coef = "ons_poak:coda_vc1"),
  prior(normal(0, 20),     class = sd),
  prior(normal(0, 25),     class = sigma),
  prior(lkj(2), class = cor)
)
```

## 4.3 Fit the models

Each equation is fit twice, as with the primary VOT model: once with
`brms`' default priors, once with the tiered informed priors above.
These are now single-equation maximal models — the same structure that
converged cleanly and reasonably quickly for the primary VOT model —
so standard settings (`adapt_delta = 0.95`, default `max_treedepth`)
should suffice; the earlier joint model's need for far more aggressive
settings was itself a symptom of that model being impractically large.

```{r mediation-fit, results='hide'}
# Mediator equation (vdur ~ ons_poa * coda_vc)
model_M_default <- brm(
  bf_vdur, data = data_mod, family = gaussian(),
  chains = 4, iter = 4000, warmup = 1500, cores = 4, seed = 123,
  control = list(adapt_delta = 0.95)
)

model_M_informed <- brm(
  bf_vdur, data = data_mod, family = gaussian(),
  prior = vdur_priors,
  chains = 4, iter = 4000, warmup = 1500, cores = 4, seed = 123,
  control = list(adapt_delta = 0.95)
)

# Outcome equation (vot ~ vdur + ons_poa * coda_vc)
model_Y_default <- brm(
  bf_vot, data = data_mod, family = gaussian(),
  chains = 4, iter = 4000, warmup = 1500, cores = 4, seed = 123,
  control = list(adapt_delta = 0.95)
)

model_Y_informed <- brm(
  bf_vot, data = data_mod, family = gaussian(),
  prior = vot_priors,
  chains = 4, iter = 4000, warmup = 1500, cores = 4, seed = 123,
  control = list(adapt_delta = 0.95)
)

# Save immediately -- these take a while to fit and losing them to a
# crashed/restarted session is costly
saveRDS(model_M_default,  "model_M_default.rds")
saveRDS(model_M_informed, "model_M_informed.rds")
saveRDS(model_Y_default,  "model_Y_default.rds")
saveRDS(model_Y_informed, "model_Y_informed.rds")
```

## 4.4 Convergence diagnostics

```{r mediation-diagnostics}
print(summary(model_M_informed), digits = 2)
sum(rstan::get_divergent_iterations(model_M_informed$fit))

print(summary(model_Y_informed), digits = 2)
sum(rstan::get_divergent_iterations(model_Y_informed$fit))
```

## 4.5 Compute indirect, direct, and total effects

Path *a* comes from the mediator model's `coda_vc1` coefficient; path
*b* and path *c* both come from the outcome model (`vdur` and
`coda_vc1` coefficients, respectively). Because *a* and *b* come from
two separately-fit models, we approximate their joint distribution by
pairing posterior draws index-by-index (both models are fit with the
same number of post-warmup draws per chain and the same seed
structure, so this pairing is arbitrary but consistent -- it does
*not* reflect genuine posterior correlation between the two
equations' parameters, only their marginal uncertainty).

```{r mediation-effects}
summarize_draws_custom <- function(x) {
  c(estimate = mean(x),
    lower    = as.numeric(quantile(x, 0.025)),
    upper    = as.numeric(quantile(x, 0.975)))
}

compute_mediation_separate <- function(model_M, model_Y) {
  a <- as_draws_df(model_M)$b_coda_vc1
  b <- as_draws_df(model_Y)$b_vdur
  c_direct <- as_draws_df(model_Y)$b_coda_vc1

  indirect <- a * b               # approximate: a, b treated as independent
  direct   <- c_direct
  total    <- direct + indirect
  prop_mediated <- indirect / total

  list(
    a             = summarize_draws_custom(a),
    b             = summarize_draws_custom(b),
    indirect      = summarize_draws_custom(indirect),
    direct        = summarize_draws_custom(direct),
    total         = summarize_draws_custom(total),
    prop_mediated = summarize_draws_custom(prop_mediated),
    draws = list(a = a, b = b, indirect = indirect,
                 direct = direct, total = total)
  )
}

med_informed <- compute_mediation_separate(model_M_informed, model_Y_informed)
med_default  <- compute_mediation_separate(model_M_default,  model_Y_default)

bind_rows(
  a             = med_informed$a,
  b             = med_informed$b,
  indirect      = med_informed$indirect,
  direct        = med_informed$direct,
  total         = med_informed$total,
  prop_mediated = med_informed$prop_mediated,
  .id = "quantity"
) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))
```

## 4.6 Default vs. informed priors

```{r mediation-compare}
loo_M_default  <- loo(model_M_default)
loo_M_informed <- loo(model_M_informed)
loo_compare(loo_M_default, loo_M_informed)

loo_Y_default  <- loo(model_Y_default)
loo_Y_informed <- loo(model_Y_informed)
loo_compare(loo_Y_default, loo_Y_informed)

# Direct comparison of the indirect effect under both prior specs
summarize_draws_custom(med_default$draws$indirect)
summarize_draws_custom(med_informed$draws$indirect)
```

## 4.7 Visualize the posterior distributions

```{r mediation-plot, fig.width=8, fig.height=5}
plot_df <- tibble(
  effect = rep(c("Indirect (a x b)", "Direct (c)", "Total"), each = length(med_informed$draws$indirect)),
  value  = c(med_informed$draws$indirect, med_informed$draws$direct, med_informed$draws$total)
) %>%
  mutate(effect = factor(effect, levels = c("Total", "Direct (c)", "Indirect (a x b)")))

ggplot(plot_df, aes(x = value, y = effect, fill = effect)) +
  ggridges::geom_density_ridges(alpha = 0.7, scale = 0.9) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  labs(x = "Effect on onset VOT (ms)", y = "",
       title = "Posterior distributions: indirect, direct, and total effects") +
  theme_minimal() +
  theme(legend.position = "none")
```

---

# 5. The fully joint multivariate model (attempted, not primary)

For completeness and reproducibility, this is the fully joint model
that was attempted before settling on the two-separate-equations
approach above. It links both equations' random effects into a single
correlation matrix via `brms`' `|q|` grouping syntax, which would in
principle be the more complete analysis (it does not assume
independence between a speaker's mediator- and outcome-equation random
effects, unlike Section 4's approach). In practice, this model proved
impractically slow to sample even with `adapt_delta = 0.99` and
`max_treedepth = 15`, and was abandoned in favor of Section 4's
approach. It may be worth revisiting with substantially more
compute time, a longer allowed runtime, or a reduced random-slopes
structure (e.g., dropping place-of-articulation random slopes while
keeping the focal `coda_vc` and `vdur` slopes) if a fully joint
estimate is needed for a future revision.

```{r mediation-joint, eval=FALSE}
bf_vdur_joint <- bf(vdur ~ ons_poa * coda_vc + (1 + ons_poa * coda_vc | q | sub))
bf_vot_joint  <- bf(vot  ~ vdur + ons_poa * coda_vc + (1 + vdur + ons_poa * coda_vc | q | sub))

mediation_informed_priors_joint <- c(
  prior(normal(0, 150), class = Intercept, resp = vdur, lb = 0),
  prior(normal(55, 30), class = b, coef = coda_vc1, resp = vdur),
  prior(normal(0, 25),  class = b, coef = "ons_poat:coda_vc1", resp = vdur),
  prior(normal(0, 25),  class = b, coef = "ons_poak:coda_vc1", resp = vdur),
  prior(normal(0, 30),  class = sd,    resp = vdur),
  prior(normal(0, 30),  class = sigma, resp = vdur),

  prior(normal(0, 100),     class = Intercept, resp = vot, lb = 0),
  prior(normal(15, 8),      class = b, coef = ons_poat, resp = vot),
  prior(normal(30, 12),     class = b, coef = ons_poak, resp = vot),
  prior(normal(0.15, 0.15), class = b, coef = vdur, resp = vot),
  prior(normal(0, 20),      class = b, coef = coda_vc1, resp = vot),
  prior(normal(0, 15),      class = b, coef = "ons_poat:coda_vc1", resp = vot),
  prior(normal(0, 15),      class = b, coef = "ons_poak:coda_vc1", resp = vot),
  prior(normal(0, 20),      class = sd,    resp = vot),
  prior(normal(0, 25),      class = sigma, resp = vot),

  prior(lkj(2), class = cor)
)

bayes_med_joint <- brm(
  bf_vdur_joint + bf_vot_joint + set_rescor(FALSE),
  data    = data_mod,
  family  = gaussian(),
  prior   = mediation_informed_priors_joint,
  chains  = 4,
  iter    = 4000,
  warmup  = 1500,
  cores   = 4,
  seed    = 123,
  control = list(adapt_delta = 0.99, max_treedepth = 15)
)
```

---

# 6. Summary

The two-equation Bayesian mediation analysis in Section 4 provides
estimates of the indirect effect of coda voicing on onset VOT,
operating through vowel duration (path $a \times b$), and the direct,
residual effect of coda voicing on onset VOT after controlling for
vowel duration (path $c$), each with full posterior uncertainty from
a maximal random-effects structure — without relying on the
frequentist bootstrap or random-intercept-only assumptions of the
original SEM/`mediation` approach. The one respect in which this
falls short of a fully consistent maximal-structure analysis is that
the two equations' random effects are treated as independent, since
the fully joint model (Section 5) was not tractable in practice.

---

# 7. Appendix: Frequentist comparison (optional robustness check)

The code below reproduces the original `lavaan`/`mediation` approach
for comparison purposes only. It is **not** the primary analysis: it
uses random-intercept-only mixed models and bootstrap/quasi-Bayesian
inference rather than the maximal-structure models above, and (once
corrected to ms, as below) its coefficients are on a different fitted
scale than the primary model's. Treat any correspondence between the
two as a rough sanity check, not confirmatory evidence.

```{r frequentist-appendix, eval=FALSE}
library(lavaan)
library(lme4)
library(mediation)

# Dummy-coded POA for lavaan (does not handle factors the way lme4 does)
data_mod$alveolar <- ifelse(data_mod$ons_poa == "t", 1, 0)
data_mod$velar    <- ifelse(data_mod$ons_poa == "k", 1, 0)
data_mod$coda_vc_num <- as.numeric(as.character(data_mod$coda_vc))

sem_model <- '
  vdur ~ a*coda_vc_num + alveolar + velar
  vot  ~ b*vdur + c*coda_vc_num + alveolar + velar
  indirect := a * b
  direct   := c
  total    := c + (a * b)
'

fit_sem <- sem(sem_model, data = data_mod, se = "bootstrap",
                bootstrap = 5000, cluster = "sub")
summary(fit_sem, fit.measures = TRUE)

model_M <- lmer(vdur ~ coda_vc + ons_poa + (1 | sub), data = data_mod, REML = FALSE)
model_Y <- lmer(vot  ~ vdur + coda_vc + ons_poa + (1 | sub), data = data_mod, REML = FALSE)

med_out <- mediate(model_M, model_Y, treat = "coda_vc", mediator = "vdur",
                    boot = FALSE, sims = 1000)
summary(med_out)
```
