---
title: "Direct and Indirect Effects of Coda Voicing on Onset VOT"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
# This chunk sets global options for all subsequent chunks.
# include=FALSE means this chunk runs but produces no output in the knitted file.
knitr::opts_chunk$set(
  echo = TRUE,        # Show R code in the output
  warning = FALSE,    # Suppress warnings
  message = FALSE     # Suppress package loading messages
)
```

---

# Overview

This analysis examines the **voicing effect** in English monosyllables: the well-documented
phenomenon whereby voiced codas are preceded by longer vowels than voiceless codas. 
We extend this by asking whether coda voicing also exerts a **long-distance, indirect effect** 
on onset Voice Onset Time (VOT), mediated by vowel duration.

The hypothesized causal chain is:

$$\text{Coda Voicing} \xrightarrow{a} \text{Vowel Duration} \xrightarrow{b} \text{VOT}$$

We use **Structural Equation Modelling (SEM)** via the `lavaan` package to estimate:

- The **direct effect** of coda voicing on vowel duration (path *a*)
- The **direct effect** of vowel duration on VOT (path *b*)
- The **indirect effect** of coda voicing on VOT via vowel duration (path *a* × *b*)
- The **direct effect** of coda voicing on VOT (path *c*, bypassing vowel duration)

Because speakers vary substantially, we also fit a parallel **mixed-effects mediation 
analysis** using the `mediation` package with `lme4` models, which explicitly accounts 
for by-speaker random effects.

---

# 1. Load Packages

```{r load-packages}
# Install any packages you do not yet have by uncommenting the line below:
# install.packages(c("lavaan", "lme4", "mediation", "semPlot", "tidyverse"))

library(lavaan)      # Structural equation modelling
library(lme4)        # Mixed-effects models (for the mediation approach)
library(mediation)   # Mediation analysis with mixed-effects models
library(semPlot)     # Path diagram visualization for SEM
library(tidyverse)   # Data manipulation and plotting
```

---

# 2. Prepare the Data

```{r prepare-data}
# -----------------------------------------------------------------------
# IMPORTANT: Replace this section with your own data loading step.
# For example: data <- read.csv("your_data.csv")
# The analysis below assumes a dataframe called "data" with the 
# following columns:
#   sub      : speaker identifier (factor)
#   ons_poa  : onset place of articulation (1=bilabial, 2=alveolar, 3=velar)
#   coda_vc  : coda voicing (0=voiceless, 1=voiced)
#   vdur     : vowel duration (numeric, milliseconds)
#   vot      : voice onset time (numeric, milliseconds)
# -----------------------------------------------------------------------

data <- read.csv("/Users/chandan/Desktop/data_12-11-24.csv", header=TRUE)


# Convert speaker ID to a factor if it is not already.
# This ensures R treats it as a grouping variable, not a number.
data$sub <- as.factor(data$sub)

# Convert onset place of articulation to a labelled factor.
# This prevents R from treating it as a continuous numeric variable,
# which would be phonologically meaningless (velar is not "more" than bilabial).
data$ons_poa <- factor(data$ons_poa,
                       levels = c(1, 2, 3),
                       labels = c("bilabial", "alveolar", "velar"))

# Coda voicing is already coded 0/1, which is what we need.
# 0 = voiceless coda, 1 = voiced coda.
# No transformation needed, but make sure R sees it as numeric.
data$coda_vc <- as.numeric(data$coda_vc)

# Create dummy variables for onset place of articulation.
# lavaan does not handle factors automatically the way lme4 does,
# so we need to create these explicitly.
# Bilabial (level 1) serves as the reference category.
data$alveolar <- ifelse(data$ons_poa == "alveolar", 1, 0)
data$velar    <- ifelse(data$ons_poa == "velar",    1, 0)

# Quick check: make sure the data looks as expected
glimpse(data)
```

---

# 3. Descriptive Statistics

```{r descriptives}
# It is always good practice to examine your data before modelling.
# Here we look at mean VOT and vowel duration by coda voicing and place.

data %>%
  group_by(ons_poa, coda_vc) %>%
  summarise(
    n           = n(),
    mean_vot    = round(mean(vot,  na.rm = TRUE), 2),
    sd_vot      = round(sd(vot,    na.rm = TRUE), 2),
    mean_vdur   = round(mean(vdur, na.rm = TRUE), 2),
    sd_vdur     = round(sd(vdur,   na.rm = TRUE), 2),
    .groups = "drop"
  )
```

```{r plot-descriptives, fig.width=10, fig.height=4}
# Visualize the key relationships before modelling.
# Panel 1: Vowel duration by coda voicing (the classic voicing effect)
# Panel 2: VOT by coda voicing (the long-distance effect we are testing)

p1 <- ggplot(data, aes(x = factor(coda_vc, labels = c("Voiceless", "Voiced")),
                        y = vdur, fill = factor(coda_vc))) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~ons_poa) +
  labs(title = "Vowel Duration by Coda Voicing and Onset Place",
       x = "Coda Voicing", y = "Vowel Duration (ms)") +
  theme_minimal() +
  theme(legend.position = "none")

p2 <- ggplot(data, aes(x = factor(coda_vc, labels = c("Voiceless", "Voiced")),
                        y = vot, fill = factor(coda_vc))) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~ons_poa) +
  labs(title = "VOT by Coda Voicing and Onset Place",
       x = "Coda Voicing", y = "VOT (ms)") +
  theme_minimal() +
  theme(legend.position = "none")

# Print both plots side by side
gridExtra::grid.arrange(p1, p2, ncol = 2)
```

---

# 4. Structural Equation Model (lavaan)

The SEM approach estimates all paths simultaneously in a single model.
It does not natively handle random effects, but we include cluster-robust
standard errors to partially account for speaker clustering.

## 4.1 Specify the Model

```{r sem-model-spec}
# In lavaan syntax:
#   ~   means "is regressed on"
#   *   is used to label a path coefficient (e.g., a*coda_vc labels the 
#       coefficient of coda_vc as "a" so we can refer to it later)
#   :=  defines a new parameter as a function of existing ones

model <- '
  # ---------------------------------------------------------
  # MEDIATOR MODEL
  # Coda voicing predicts vowel duration (path a).
  # We also control for onset place of articulation.
  # ---------------------------------------------------------
  vdur ~ a*coda_vc + alveolar + velar

  # ---------------------------------------------------------
  # OUTCOME MODEL
  # Vowel duration predicts VOT (path b).
  # Coda voicing also has a potential direct path to VOT (path c).
  # We again control for onset place.
  # ---------------------------------------------------------
  vot ~ b*vdur + c*coda_vc + alveolar + velar

  # ---------------------------------------------------------
  # DEFINED PARAMETERS
  # These compute the effects we care about from the labeled paths above.
  # ---------------------------------------------------------
  
  # Indirect effect: how much of the coda voicing effect on VOT
  # is carried through vowel duration? This is the key quantity.
  indirect := a * b

  # Direct effect: the remaining effect of coda voicing on VOT
  # after accounting for the mediation through vowel duration.
  direct := c

  # Total effect: the sum of direct and indirect paths.
  total := c + (a * b)
'
```

## 4.2 Fit the Model

```{r sem-fit}
# We use bootstrap standard errors (1000 resamples) because the indirect 
# effect (a product of two coefficients) is not normally distributed.
# Bootstrap confidence intervals are more reliable for inference on 
# indirect effects than standard asymptotic SEs.
#
# se = "bootstrap" requests bootstrapped SEs.
# bootstrap = 1000 sets the number of resamples (increase to 5000 for publication).
# cluster = "sub" provides cluster-robust resampling by speaker,
# which partially addresses the non-independence of observations.

set.seed(42)  # Set seed for reproducibility

fit <- sem(model,
           data      = data,
           se        = "bootstrap",
           bootstrap = 5000,
           cluster   = "sub")
```

## 4.3 Examine Model Fit

```{r sem-fit-indices}
# Before interpreting path coefficients, check whether the model fits the data well.
# Key fit indices and conventional benchmarks:
#   CFI  > 0.95  (Comparative Fit Index; higher is better)
#   RMSEA < 0.06 (Root Mean Square Error of Approximation; lower is better)
#   SRMR  < 0.08 (Standardized Root Mean Square Residual; lower is better)

summary(fit, fit.measures = TRUE)
```

## 4.4 Extract Parameter Estimates

```{r sem-parameters}
# Extract the full parameter table with bootstrap confidence intervals.
# boot.ci.type = "bca" requests bias-corrected accelerated intervals,
# which are more accurate than percentile intervals for indirect effects.

parameterEstimates(fit,
                   ci            = TRUE,
                   boot.ci.type  = "bca.simple",
                   standardized  = TRUE) %>%
  filter(op %in% c("~", ":=")) %>%   # Keep regression paths and defined parameters
  select(lhs, op, rhs, est, se, ci.lower, ci.upper, pvalue, std.all) %>%
  mutate(across(where(is.numeric), ~round(.x, 4)))
```

## 4.5 Path Diagram

```{r sem-path-diagram, fig.width=8, fig.height=5}
# semPlot produces a visual representation of the path model.
# This is useful for presentations and papers.

semPaths(fit,
         what       = "std",          # Show standardized coefficients
         whatLabels = "std",
         style      = "ram",
         layout     = "tree2",
         residuals  = FALSE,
         intercepts = FALSE,
         nCharNodes = 8,
         edge.label.cex = 0.9,
         title      = FALSE)

title("SEM Path Diagram: Indirect Effect of Coda Voicing on VOT\n(standardized coefficients)")
```

---

# 5. Mixed-Effects Mediation Analysis (lme4 + mediation)

The `mediation` package fits the two component models as mixed-effects
models with `lme4`, which properly accounts for by-speaker random effects.
This is the more conservative and statistically appropriate approach given
the speaker variation in your data.

## 5.1 Fit the Two Component Models

```{r mediation-models}
# MODEL 1: The mediator model
# Does coda voicing predict vowel duration, after accounting for
# onset place and by-speaker variability?
# (1 | sub) adds a random intercept for each speaker.

model.M <- lmer(vdur ~ coda_vc + ons_poa +
                  (1 | sub),
                data = data,
                REML = FALSE)   # REML = FALSE is required for model comparison

# MODEL 2: The outcome model
# Does vowel duration predict VOT, after controlling for coda voicing,
# onset place, and by-speaker variability?

model.Y <- lmer(vot ~ vdur + coda_vc + ons_poa +
                  (1 | sub),
                data = data,
                REML = FALSE)

# Quick summaries to check the component models before running mediation
summary(model.M)
summary(model.Y)
```

## 5.2 Run the Mediation Analysis

```{r mediation-run}
# mediate() takes the two fitted models and estimates the indirect effect
# using quasi-Bayesian Monte Carlo simulation (or bootstrapping if boot = TRUE).
#
# treat     = the independent variable (coda voicing)
# mediator  = the mediating variable (vowel duration)
# boot      = TRUE requests bootstrapped confidence intervals
# boot.ci.type = "bca" requests bias-corrected accelerated intervals
# sims      = number of simulations (increase to 5000 for publication)

set.seed(42)

med.out <- mediate(model.M,
                   model.Y,
                   treat        = "coda_vc",
                   mediator     = "vdur",
                   boot         = FALSE,
                   boot.ci.type = "bca",
                   sims         = 1000)

summary(med.out)
```

## 5.3 Interpret the Output

The `mediate()` summary reports four key quantities:

| Term | Meaning |
|------|---------|
| **ACME** | Average Causal Mediation Effect — the indirect effect of coda voicing on VOT *via* vowel duration. This is the key result. |
| **ADE** | Average Direct Effect — the effect of coda voicing on VOT after controlling for vowel duration. |
| **Total Effect** | ACME + ADE — the overall effect of coda voicing on VOT. |
| **Prop. Mediated** | The proportion of the total effect carried by the indirect path. |

If the **95% CI for ACME excludes zero**, we have reliable evidence for an
indirect effect of coda voicing on VOT mediated by vowel duration.

## 5.4 Visualize the Mediation Results

```{r mediation-plot, fig.width=7, fig.height=4}
# The default plot shows ACME, ADE, and Total Effect with confidence intervals.
plot(med.out,
     main = "Mediation Analysis: Indirect Effect of Coda Voicing on VOT")
```

---

# 6. Expanding the Random Effects Structure

The models above use random intercepts only. If you want a more conservative
and thorough analysis, you can add random slopes, allowing the effects of
interest to vary across speakers. This is theoretically motivated because
speakers likely differ not just in their baseline VOT and vowel duration,
but in how strongly their vowel duration responds to coda voicing.

```{r random-slopes, eval=FALSE}
# NOTE: eval=FALSE means this chunk is shown but not run automatically.
# Run it manually if your data support it. Random slopes models 
# can fail to converge with small datasets or few speakers.

model.M.slopes <- lmer(vdur ~ coda_vc + ons_poa +
                         (1 + coda_vc | sub),   # Random slope for coda voicing
                       data    = data,
                       REML    = FALSE,
                       control = lmerControl(optimizer = "bobyqa"))

model.Y.slopes <- lmer(vot ~ vdur + coda_vc + ons_poa +
                         (1 + vdur + coda_vc | sub),  # Random slopes for both predictors
                       data    = data,
                       REML    = FALSE,
                       control = lmerControl(optimizer = "bobyqa"))

# If these converge, re-run the mediation with the slope models:
set.seed(42)
med.out.slopes <- mediate(model.M.slopes,
                           model.Y.slopes,
                           treat        = "coda_vc",
                           mediator     = "vdur",
                           boot         = FALSE,
                           boot.ci.type = "bca",
                           sims         = 1000)
summary(med.out.slopes)
```

---

# 7. Summary of Results

```{r results-summary}
# Pull the key numbers from the mediation output for a clean summary table.

med_summary <- data.frame(
  Effect            = c("Indirect (ACME)", "Direct (ADE)", "Total", "Proportion Mediated"),
  Estimate          = c(med.out$d.avg,    med.out$z.avg,
                        med.out$tau.coef, med.out$n.avg),
  CI_Lower          = c(med.out$d.avg.ci[1], med.out$z.avg.ci[1],
                        med.out$tau.ci[1],   med.out$n.avg.ci[1]),
  CI_Upper          = c(med.out$d.avg.ci[2], med.out$z.avg.ci[2],
                        med.out$tau.ci[2],   med.out$n.avg.ci[2]),
  p_value           = c(med.out$d.avg.p,  med.out$z.avg.p,
                        med.out$tau.p,    med.out$n.avg.p)
) %>%
  mutate(across(where(is.numeric), ~round(.x, 4)))

med_summary
```

---

# 8. Session Info

```{r session-info}
# Always include session info in reproducible analyses so readers know
# exactly which versions of R and each package were used.
sessionInfo()
```
