---
title: "Burst Voice: Onset VOT ~ Place of Articulation x Coda Voicing"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Data analysis

## Setup and data preparation

`ons_poa` is coded once here as a factor with levels/labels
`c("p", "t", "k")`, and `coda_vc` once as a factor with levels
`c(0, 1)`. Every plot, model, and prediction grid in the rest of this
document inherits this coding and reference level (`/p/`), rather than
each section re-deriving place-of-articulation labels independently.

```{r libraries}
library(ggplot2)
library(tidyverse)
library(dplyr)
library(ggridges)
library(lme4)
library(brms)
library(marginaleffects)
library(knitr)
```

```{r data-prep}
data <- read.csv("/Users/chandan/Desktop/GitHub/Burst_voice/Data/data_12-11-24.csv",
                  header = TRUE)

data <- data %>%
  mutate(
    ons_poa = factor(ons_poa, levels = c(1, 2, 3), labels = c("p", "t", "k")),
    coda_vc = factor(coda_vc, levels = c(0, 1)),
    # Convert VOT and vowel duration from seconds (as stored in the
    # raw CSV) to milliseconds. This matters beyond readability: the
    # informed priors specified later (e.g., normal(30, 10) for the
    # /p/ intercept) were written assuming a millisecond-scale
    # outcome, but were previously being applied to vot/vdur while
    # still in seconds -- a units mismatch that made those priors
    # far less informative than intended (a prior centered at "30"
    # is only meaningfully informative if the outcome is actually on
    # a scale where 30 is a plausible value). Converting here keeps
    # the data and the priors on the same scale throughout.
    vot  = vot * 1000,
    vdur = vdur * 1000,
    vot_ratio = vot / vdur,   # unitless; unaffected by the ms conversion
    CVdur = vot + vdur
  )

# single data frame used throughout; kept as a separate name only
# for readability in code carried over from earlier drafts
data_mod <- data

number_of_subs <- length(unique(data$sub))
number_of_subs
```

## Descriptive statistics

Basic descriptive statistics for onset VOT, vowel duration, the
VOT/vowel-duration ratio, and combined CV duration, broken down by
place of articulation and coda voicing (the two manipulated factors)
and reported alongside the overall totals. `sd` is the standard
deviation of individual tokens within each cell (not a standard
error), so these numbers reflect the same raw variability visualized
in the boxplots and ridge plots below, before any model-based
smoothing.

```{r descriptive-stats-by-condition}
descriptives_by_condition <- data %>%
  group_by(ons_poa, coda_vc, vowel) %>%
  summarise(
    n            = n(),
    mean_vot     = mean(vot, na.rm = TRUE),
    sd_vot       = sd(vot, na.rm = TRUE),
    mean_vdur    = mean(vdur, na.rm = TRUE),
    sd_vdur      = sd(vdur, na.rm = TRUE),
    mean_ratio   = mean(vot_ratio, na.rm = TRUE),
    sd_ratio     = sd(vot_ratio, na.rm = TRUE),
    mean_CVdur   = mean(CVdur, na.rm = TRUE),
    sd_CVdur     = sd(CVdur, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(across(where(is.numeric), ~ round(.x, 2))) %>%
  arrange(ons_poa, coda_vc, vowel) %>%
  rename(
    `Onset POA`   = ons_poa,
    `Coda voice`  = coda_vc,
    `Vowel`       = vowel,
    `N`           = n,
    `Mean VOT`    = mean_vot,
    `SD VOT`      = sd_vot,
    `Mean Vdur`   = mean_vdur,
    `SD Vdur`     = sd_vdur,
    `Mean ratio`  = mean_ratio,
    `SD ratio`    = sd_ratio,
    `Mean CVdur`  = mean_CVdur,
    `SD CVdur`    = sd_CVdur
  )

kable(descriptives_by_condition,
      caption = "Descriptive statistics by onset place of articulation, coda voicing, and vowel identity. VOT, Vdur, and CVdur are in ms; ratio (VOT/Vdur) is unitless.")
```

For reference, the same statistics collapsed across place of
articulation and coda voicing (i.e., the marginal/overall values):

```{r descriptive-stats-overall}
descriptives_overall <- data %>%
  summarise(
    n            = n(),
    mean_vot     = mean(vot, na.rm = TRUE),
    sd_vot       = sd(vot, na.rm = TRUE),
    mean_vdur    = mean(vdur, na.rm = TRUE),
    sd_vdur      = sd(vdur, na.rm = TRUE),
    mean_ratio   = mean(vot_ratio, na.rm = TRUE),
    sd_ratio     = sd(vot_ratio, na.rm = TRUE),
    mean_CVdur   = mean(CVdur, na.rm = TRUE),
    sd_CVdur     = sd(CVdur, na.rm = TRUE)
  ) %>%
  mutate(across(everything(), ~ round(.x, 2))) %>%
  rename(
    `N`           = n,
    `Mean VOT`    = mean_vot,
    `SD VOT`      = sd_vot,
    `Mean Vdur`   = mean_vdur,
    `SD Vdur`     = sd_vdur,
    `Mean ratio`  = mean_ratio,
    `SD ratio`    = sd_ratio,
    `Mean CVdur`  = mean_CVdur,
    `SD CVdur`    = sd_CVdur
  )

kable(descriptives_overall,
      caption = "Overall descriptive statistics, collapsed across place of articulation and coda voicing.")
```

Descriptive statistics by subject (mean VOT and vowel duration per
speaker, collapsed across condition), useful for spotting any speaker
whose overall productions look unusually fast/slow or short/long
relative to the group before interpreting the by-subject random
effects in the Bayesian models below:

```{r descriptive-stats-by-subject}
descriptives_by_subject <- data %>%
  group_by(sub) %>%
  summarise(
    n         = n(),
    mean_vot  = mean(vot, na.rm = TRUE),
    sd_vot    = sd(vot, na.rm = TRUE),
    mean_vdur = mean(vdur, na.rm = TRUE),
    sd_vdur   = sd(vdur, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(across(where(is.numeric), ~ round(.x, 2))) %>%
  arrange(sub)

kable(descriptives_by_subject,
      caption = "Descriptive statistics by subject, collapsed across condition.")
```

## Preliminary plotting

### By onset POA

This is a first pass at looking at the data.

```{r consistency-check}
# Within-condition consistency (coefficient of variation) check
consistency_stats <- data %>%
  group_by(sub, ons_poa, coda_vc, vowel) %>%
  summarise(
    n_obs = n(),
    mean_CVdur = mean(CVdur, na.rm = TRUE),
    sd_CVdur = sd(CVdur, na.rm = TRUE),
    cv = sd_CVdur / mean_CVdur * 100,
    .groups = "drop"
  ) %>%
  filter(n_obs > 1)

consistency_stats %>%
  arrange(desc(cv)) %>%
  head(20)
```

```{r summary-table}
# Summary table of mean VOT by place and coda voicing
summary_table <- data %>%
  group_by(ons_poa, coda_vc) %>%
  summarize(
    mean_vot = mean(vot, na.rm = TRUE),
    vot_range = paste0(min(vot, na.rm = TRUE), " to ", max(vot, na.rm = TRUE))
  ) %>%
  ungroup()

print(summary_table)
```

```{r plot-overall, fig.width=8, fig.height=5}
# Boxplot: VOT by onset POA and coda voicing
plot_overall <- ggplot(data, aes(x = ons_poa, y = vot, fill = coda_vc)) +
  geom_boxplot(notch = TRUE, outlier.shape = NA) +
  theme(axis.text.x = element_blank()) +
  labs(x = NULL, y = "CV duration (ms)") +
  facet_wrap(~ons_poa, scales = "free") +
  scale_y_continuous(limits = c(15, 180)) +
  scale_fill_discrete(name = "Coda voice", labels = c("-vc", "+vc"))

print(plot_overall)
```

```{r plot-overall-by-sub, fig.width=10, fig.height=8}
plot_overall + facet_wrap(~sub)
```

```{r plot-cvdur, fig.width=8, fig.height=5}
# Boxplot: CV duration by onset POA and coda voicing
plot_overall_CVdur <- ggplot(data, aes(x = ons_poa, y = CVdur, fill = coda_vc)) +
  geom_boxplot(notch = TRUE, outlier.shape = NA) +
  theme(axis.text.x = element_blank()) +
  labs(x = NULL, y = "CV duration (ms)") +
  facet_wrap(~ons_poa, scales = "free") +
  scale_fill_discrete(name = "Coda voice", labels = c("-vc", "+vc"))

print(plot_overall_CVdur)
```

```{r plot-vot-poa, fig.width=10, fig.height=8}
# Boxplot: VOT by onset POA, faceted by subject
plot_vot_poa <- ggplot(data, aes(x = vot, y = ons_poa, fill = coda_vc)) +
  geom_boxplot(notch = TRUE, outlier.shape = NA, width = 0.5) +
  theme_minimal() +
  labs(x = "VOT (ms)", y = "Onset POA", fill = "Coda voice") +
  scale_fill_discrete(labels = c("[-vc]", "[+vc]"))

print(plot_vot_poa + facet_wrap(~sub))
```

```{r plot-vdur-poa, fig.width=8, fig.height=5}
# Boxplot: vowel duration by onset POA, faceted by subject
plot_vdur_poa <- ggplot(data, aes(x = vdur, y = ons_poa, fill = coda_vc)) +
  geom_boxplot(notch = TRUE, outlier.shape = NA, width = 0.5) +
  theme_minimal() +
  labs(x = "Vdur (ms)", y = "Onset POA", fill = "Coda voice") +
  scale_x_continuous(limits = c(0, 300), expand = c(0, 0)) +
  scale_fill_discrete(labels = c("[-vc]", "[+vc]"))

plot_vdur_poa
```

```{r plot-vdur-poa-by-sub, fig.width=10, fig.height=8}
plot_vdur_poa + facet_wrap(~sub)
```

### By onset POA and vowel

Ridge plots (stacked density plots) for voiceless and voiced coda
tokens by POA and vowel.

```{r plot-vowel-boxplot, fig.width=9, fig.height=6}
plot_vowel <- ggplot(data, aes(x = vowel, y = vot, fill = coda_vc)) +
  geom_boxplot(notch = TRUE) +
  labs(y = "VOT (ms)", x = "Vowel", fill = "Coda voicing") +
  facet_wrap(~ons_poa, scale = "free")
plot_vowel
```

```{r ridge-labels}
# ons_poa is already factor-coded p/t/k, so these labellers map from
# that coding to display symbols only
poa_labels <- c("p" = "/p/", "t" = "/t/", "k" = "/k/")
vowel_labels <- c("ae" = "\u00e6", "aw" = "\u0254", "eh" = "\u025b",
                   "ih" = "\u026a", "uh" = "\u028c")
```

```{r ridge-by-vowel, fig.width=12, fig.height=6}
plot_vowel_ridges <- ggplot(data, aes(x = vot, y = coda_vc, fill = coda_vc)) +
  geom_density_ridges(alpha = 0.7) +
  scale_x_continuous(breaks = c(0, 100, 200), limits = c(0, NA)) +
  scale_y_discrete(labels = c("0" = "Voiceless", "1" = "Voiced")) +
  scale_fill_manual(
    name = "Coda",
    values = c("0" = "#F8766D", "1" = "#00BFC4"),
    labels = c("0" = "Voiceless", "1" = "Voiced")
  ) +
  labs(x = "VOT (ms)", y = "") +
  facet_grid(vowel ~ ons_poa, scales = "free_x",
             labeller = labeller(ons_poa = poa_labels, vowel = vowel_labels)) +
  theme_ridges() +
  theme(axis.text.y = element_blank(),
        axis.title.x = element_text(hjust = 0.5),
        strip.text = element_text(margin = margin(t = 4, b = 4)),
        strip.text.x = element_text(margin = margin(t = 3, b = 4)),
        strip.text.y = element_text(margin = margin(l = 3, r = 3)),
        panel.spacing = unit(0.3, "cm"))

print(plot_vowel_ridges)
ggsave("plot_ridges.png", plot = plot_vowel_ridges, width = 12, height = 6, dpi = 300)
```

```{r ridge-collapsed, fig.width=9, fig.height=6}
# Same ridge plot, collapsed across vowel
plot_ridges <- ggplot(data, aes(x = vot, y = coda_vc, fill = coda_vc)) +
  geom_density_ridges(alpha = 0.7) +
  scale_x_continuous(breaks = c(0, 50, 100, 150, 200), limits = c(0, NA)) +
  scale_y_discrete(labels = c("0" = "Voiceless", "1" = "Voiced"), expand = c(0, 0)) +
  scale_fill_manual(
    name = "Coda",
    values = c("0" = "#F8766D", "1" = "#00BFC4"),
    labels = c("0" = "Voiceless", "1" = "Voiced")
  ) +
  labs(x = "VOT (ms)", y = "") +
  facet_wrap(~ons_poa, scales = "free_x",
             labeller = labeller(ons_poa = poa_labels)) +
  theme_ridges() +
  theme(axis.text.y = element_blank(),
        axis.title.x = element_text(hjust = 0.5),
        strip.text = element_text(margin = margin(t = 4, b = 4)),
        strip.text.x = element_text(margin = margin(t = 3, b = 4)),
        panel.spacing = unit(0.3, "cm"),
        legend.position = c(0.85, 0.85),
        legend.background = element_rect(fill = "white", color = "black", linewidth = 0.3),
        legend.margin = margin(t = 5, r = 5, b = 5, l = 5),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 10))

plot_ridges
```

```{r density-by-poa-vowel, fig.width=12, fig.height=9}
# Overlapping density plots by POA and vowel
plot_vowel_density <- ggplot(data, aes(x = vot, fill = coda_vc)) +
  geom_density(alpha = 0.6) +
  labs(x = "VOT (ms)", y = "Density", fill = "Coda voicing") +
  facet_wrap(~ons_poa + vowel, scales = "free")
plot_vowel_density
```

It looks like there is a clear effect of coda voicing on voiceless
onset VOT.

## Basic models

```{r lmer-full}
model_full <- lmer(vot ~ ons_poa * coda_vc * vowel + (1 | sub), data = data_mod)
summary(model_full)
```

Removing the vowel term (reference level `/p/`, matching the
`ons_poa` factor coding set above -- no re-recoding needed here):

```{r lmer-reduced}
model <- lmer(vot ~ ons_poa * coda_vc + (1 | sub), data = data_mod)
summary(model)
```

### Plot of the model

```{r plot-model, fig.width=9, fig.height=5}
p_lmer <- ggplot(data_mod, aes(x = vdur, y = vot, color = coda_vc)) +
  geom_point() +
  geom_smooth(method = lm) +
  xlab("Vowel dur (ms)") + ylab("VOT (ms)")
p_lmer + facet_wrap(~ons_poa)
```

### GLMM: predicting coda voicing from onset VOT and POA

```{r glmm}
model_coda <- glmer(
  coda_vc ~ vot * ons_poa + (1 | sub),
  data = data,
  family = binomial(link = "logit")
)

summary(model_coda)
```

```{r glmm-predictions, fig.width=8, fig.height=5}
# Predictions across a range of VOT values, for each POA
vot_range <- seq(from = 0, to = 250, by = 5)
sub_ref <- unique(data$sub)[1]

plot_data <- expand_grid(
  vot = vot_range,
  ons_poa = levels(data$ons_poa),   # already "p", "t", "k", in that order
  sub = sub_ref
)
plot_data$ons_poa <- factor(plot_data$ons_poa, levels = levels(data$ons_poa))
plot_data$coda_vc <- NA

plot_data$predicted_prob <- predict(model_coda, newdata = plot_data,
                                     type = "response", re.form = ~(1 | sub))

vc_pred <- ggplot(plot_data, aes(x = vot, y = predicted_prob, color = ons_poa)) +
  geom_line(linewidth = 2) +
  labs(
    title = "",
    x = "VOT (ms)",
    y = "Predicted Probability of Voiced Coda",
    color = "Onset POA"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.title = element_text(size = 14),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )

vc_pred
ggsave("vc_pred.png", plot = vc_pred, bg = "white", width = 7, height = 4, dpi = 300)
```

# Bayesian mixed-effects model of VOT ~ coda voicing

Reference level for `ons_poa` is `/p/`, set once in the data
preparation section above.

```{r bayes-prior-check}
# Confirm reference level / contrast coding before fitting
contrasts(data_mod$ons_poa)
get_prior(vot ~ ons_poa * coda_vc + (1 + ons_poa * coda_vc | sub),
          data = data_mod, family = gaussian())
```

## Maximal random effects structure (primary model)

Because onset place of articulation and coda voicing are both
within-subject manipulations, a random-intercept-only model assumes
every participant shows the same coda-voicing effect (and the same
place-of-articulation effects), differing only in their overall
baseline VOT. Following Barr, Levy, Scheepers & Tily (2013), our
primary model instead includes by-subject random slopes for place of
articulation, coda voicing, and their interaction, plus the full
correlation structure among all random effects, from the outset. This
guards against anti-conservative (too-narrow) intervals for the fixed
effects if speakers actually differ in the size of the coda-voicing
effect, not just their baseline VOT. We fit this maximal structure
first; a reduced structure is only considered as a fallback if the
maximal model fails to converge (see below).

We fit the maximal structure twice: once with `brms`' default weakly
informative priors, and once with the literature-informed priors used
throughout this document, so that the same default-vs-informed
comparison we would otherwise have run on an intercept-only model is
instead run on the model we actually report. The informed priors
below encode only what the literature actually supports well: the
*relative* place-of-articulation ordering and approximate contrast
magnitudes (/p/ $<$ /t/ $<$ /k/), and non-negativity of the absolute
/p/ baseline. They deliberately do not encode a specific absolute
baseline value for /p/, since English VOT for a given place of
articulation varies considerably across speech styles and elicitation
methods, and we have no single well-calibrated value to anchor it to.

```{r bayes-maximal-priors}
maximal_priors <- c(
  # Intercept: no informative location prior. English /p/ VOT varies
  # considerably across speech styles, elicitation methods, and
  # corpora (careful/citation-form speech tends to run longer than
  # casual speech), so we don't have a single well-calibrated
  # absolute value to anchor the baseline to -- unlike the POA
  # *contrasts* below, where the ordering and approximate relative
  # magnitude are well established across studies regardless of
  # absolute baseline. We only constrain the intercept to be
  # non-negative (VOT cannot be negative for voiceless stops in this
  # coding), via a wide, effectively flat prior over the plausible
  # range combined with a lower bound at zero.
  prior(normal(0, 100), class = Intercept, lb = 0),

  # Strongly informed: established literature on the *relative*
  # place-of-articulation ordering and approximate contrast
  # magnitude (more posterior places of articulation associated with
  # longer VOT), independent of the absolute baseline
  prior(normal(15, 8),  class = b, coef = ons_poat),
  prior(normal(30, 12), class = b, coef = ons_poak),

  # Weakly informed / regularizing: the focal, untested hypothesis
  prior(normal(0, 20),  class = b, coef = "coda_vc1"),
  prior(normal(0, 15),  class = b, coef = "ons_poat:coda_vc1"),
  prior(normal(0, 15),  class = b, coef = "ons_poak:coda_vc1"),

  # Random-effect SDs (applies to intercept + all random slopes)
  prior(normal(0, 20), class = sd),

  # Residual SD
  prior(normal(0, 25), class = sigma),

  # Correlation matrix among random effects: lkj(2) mildly shrinks
  # correlations toward zero relative to the lkj(1) (uniform) default,
  # which helps stabilize estimation with only 20 subjects
  prior(lkj(2), class = cor)
)
```

```{r bayes-maximal-fit, results='hide'}
# Maximal models are harder to sample; more iterations and a higher
# adapt_delta reduce divergent transitions at the cost of runtime.
# Both models share these settings so they're directly comparable.

# Default (weakly informative) priors, maximal random-effects structure
bayes_model_maximal_default <- brm(
  vot ~ ons_poa * coda_vc + (1 + ons_poa * coda_vc | sub),
  data    = data_mod,
  family  = gaussian(),
  chains  = 4,
  iter    = 4000,
  warmup  = 1500,
  cores   = 4,
  seed    = 123,
  control = list(adapt_delta = 0.95)
)

# Literature-informed priors, maximal random-effects structure
bayes_model_maximal_informed <- brm(
  vot ~ ons_poa * coda_vc + (1 + ons_poa * coda_vc | sub),
  data    = data_mod,
  family  = gaussian(),
  prior   = maximal_priors,
  chains  = 4,
  iter    = 4000,
  warmup  = 1500,
  cores   = 4,
  seed    = 123,
  control = list(adapt_delta = 0.95)
)
```

## Convergence diagnostics

Before interpreting either model, check for divergent transitions,
$\hat{R}$, and effective sample size. A high divergence count, $\hat{R}
> 1.02$, or very low ESS on the random-slope or correlation parameters
would indicate the data do not fully support this structure, in which
case the simplified alternatives below should be preferred over
forcing the maximal model.

```{r bayes-maximal-diagnostics}
print(summary(bayes_model_maximal_default), digits = 2)
sum(rstan::get_divergent_iterations(bayes_model_maximal_default$fit))

print(summary(bayes_model_maximal_informed), digits = 2)
sum(rstan::get_divergent_iterations(bayes_model_maximal_informed$fit))

# Visual check on chains and random-effect correlations (informed model)
plot(bayes_model_maximal_informed, variable = "^sd_", regex = TRUE)
plot(bayes_model_maximal_informed, variable = "^cor_", regex = TRUE)
```

## Default vs. informed priors, on the maximal structure

```{r bayes-maximal-loo}
loo_maximal_default  <- loo(bayes_model_maximal_default)
loo_maximal_informed <- loo(bayes_model_maximal_informed)
loo_compare(loo_maximal_default, loo_maximal_informed)
```

Direct posterior comparison for the focal coefficient:

```{r bayes-maximal-posterior-compare}
posterior_summary(bayes_model_maximal_default,  variable = "b_coda_vc1")
posterior_summary(bayes_model_maximal_informed, variable = "b_coda_vc1")
```

```{r bayes-maximal-ppcheck, fig.width=7, fig.height=5}
pp_check(bayes_model_maximal_informed, type = "hist", ndraws = 100) +
  ggtitle("Prior Predictive Check: Do simulated VOTs look reasonable?")
```

### Full-precision estimates for all fixed effects

```{r bayes-maximal-full-precision}
maximal_default_fx  <- as.data.frame(posterior_summary(bayes_model_maximal_default))  %>%
  rownames_to_column("parameter") %>%
  filter(str_starts(parameter, "b_")) %>%
  rename_with(~ paste0("default_", .x), -parameter)

maximal_informed_fx <- as.data.frame(posterior_summary(bayes_model_maximal_informed)) %>%
  rownames_to_column("parameter") %>%
  filter(str_starts(parameter, "b_")) %>%
  rename_with(~ paste0("informed_", .x), -parameter)

maximal_fixef_comparison <- left_join(maximal_default_fx, maximal_informed_fx, by = "parameter") %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))

maximal_fixef_comparison
```

## If the maximal model does not converge cleanly

Both maximal models above converged cleanly in our run (zero divergent
transitions, $\hat{R} \leq 1.00$ throughout), so no simplification was
necessary. If a future re-fit (e.g., after adding trials, changing
priors, or modifying the formula) shows divergent transitions, elevated
$\hat{R}$, or very low ESS on the random-effect correlations, simplify
in this order before abandoning the random-slopes structure entirely:

```{r bayes-maximal-simplify, eval=FALSE}
# Step 1: drop correlations among random effects, keep all slopes
# (equivalent to lme4's `||` syntax)
bayes_model_zerocorr <- brm(
  vot ~ ons_poa * coda_vc + (1 + ons_poa * coda_vc || sub),
  data    = data_mod,
  family  = gaussian(),
  prior   = maximal_priors[maximal_priors$class != "cor", ],
  chains  = 4,
  iter    = 4000,
  warmup  = 1500,
  cores   = 4,
  seed    = 123,
  control = list(adapt_delta = 0.95)
)

# Step 2: if still problematic, drop the random slope for the
# interaction term first (typically the hardest to estimate),
# retaining random slopes for the main effects
bayes_model_reduced <- brm(
  vot ~ ons_poa * coda_vc + (1 + ons_poa + coda_vc || sub),
  data    = data_mod,
  family  = gaussian(),
  prior   = maximal_priors[maximal_priors$class != "cor", ],
  chains  = 4,
  iter    = 4000,
  warmup  = 1500,
  cores   = 4,
  seed    = 123,
  control = list(adapt_delta = 0.95)
)

# Compare whichever structures were fitted; report this comparison
# alongside the final model choice
loo_compare(loo_maximal_informed, loo(bayes_model_zerocorr), loo(bayes_model_reduced))
```

## Random-intercept-only models (for comparison only)

The models below are **not** the primary reported analysis. They are
included only to document why the maximal structure above was
adopted: comparing a simpler, random-intercept-only structure against
the maximal one lets us confirm that the extra complexity is actually
supported by the data, and to show explicitly how much the
intercept-only structure understates uncertainty in the coda-voicing
effect. If you are only interested in the primary results, skip to
the Ridge Plots section.

```{r bayes-intercept-only, results='hide'}
bayes_model_default <- brm(
  vot ~ ons_poa * coda_vc + (1 | sub),
  data   = data_mod,
  family = gaussian(),
  chains = 4,
  iter   = 2000,
  cores  = 4,
  seed   = 123
)

bayes_model_informed <- brm(
  vot ~ ons_poa * coda_vc + (1 | sub),
  data   = data_mod,
  family = gaussian(),
  prior  = maximal_priors[maximal_priors$class != "cor", ],
  chains = 4,
  iter   = 2000,
  cores  = 4,
  seed   = 123
)
```

```{r bayes-intercept-only-compare}
# Does adding random slopes actually improve predictive fit?
loo_default  <- loo(bayes_model_default)
loo_informed <- loo(bayes_model_informed)
loo_compare(loo_informed, loo_maximal_informed)

# How much narrower (anti-conservative) is the intercept-only
# interval for the focal coefficient, relative to the maximal model?
posterior_summary(bayes_model_informed,          variable = "b_coda_vc1")
posterior_summary(bayes_model_maximal_informed,  variable = "b_coda_vc1")
```

## Does the VOT/vowel-duration ratio differ by coda voicing?

If onset VOT and vowel duration both lengthen together under a shared
word-level timing mechanism, the *ratio* of VOT to vowel duration
(`vot_ratio`, computed in the data-preparation step above) might stay
roughly constant across coda-voicing contexts even though the raw VOT
differs. Testing whether `vot_ratio` itself varies by coda voicing is
therefore a useful check: a credible coda-voicing effect on the ratio
would suggest VOT and vowel duration are *not* simply scaling together
proportionally, while a ratio effect indistinguishable from zero would
be consistent with proportional scaling. For consistency with the
primary VOT model above, this is fit with the same maximal
random-effects structure (by-subject random slopes for place of
articulation, coda voicing, and their interaction, plus their
correlations).

```{r bayes-ratio-model, results='hide'}
# Maximal random-effects structure, matching the main VOT model above
# for consistency. We do NOT reuse maximal_priors here: those priors
# (e.g., the POA contrast priors of normal(15,8)/normal(30,12) ms,
# the coda-voicing prior of normal(0,20) ms) are scaled for a
# millisecond-scale outcome, but vot_ratio is a unitless quantity
# with a very different scale (~0.8, roughly 0-1.5 range). Applying
# ms-scale priors here would be exactly the kind of units mismatch
# we corrected earlier for vot/vdur. brms' default priors are scaled
# automatically from the data, so we rely on those instead, with the
# same sampler settings (iterations, warmup, adapt_delta) as the main
# maximal model for consistency and comparable convergence behavior.
bayes_model_ratio <- brm(
  vot_ratio ~ ons_poa * coda_vc + (1 + ons_poa * coda_vc | sub),
  data    = data_mod,
  family  = gaussian(),
  chains  = 4,
  iter    = 4000,
  warmup  = 1500,
  cores   = 4,
  seed    = 123,
  control = list(adapt_delta = 0.95)
)
```

```{r bayes-ratio-summary}
print(summary(bayes_model_ratio), digits = 4)

# Convergence check, mirroring the main maximal model's diagnostics
sum(rstan::get_divergent_iterations(bayes_model_ratio$fit))

# Focal test: does the coda-voicing coefficient's 95% CrI exclude
# zero? If yes, the VOT/Vdur ratio differs credibly by coda voicing
# context; if the CrI spans zero, the ratio is not credibly different
# between voiced- and voiceless-coda contexts.
posterior_summary(bayes_model_ratio, variable = "b_coda_vc1")
```

# Ridge plots with posterior predictions

Ridge plots with posterior predictions overlaid, for the primary
maximal-random-effects models (default and informed priors), on the
shared `/p/` reference level.

```{r ridge-pred-setup}
data_mod$poa_coda <- factor(
  interaction(data_mod$coda_vc, data_mod$ons_poa),
  levels = c("0.p", "1.p", "0.t", "1.t", "0.k", "1.k")
)

data_mod$coda_vc_factor <- factor(
  data_mod$coda_vc,
  levels = c("1", "0"),
  labels = c("Voiced", "Voiceless")
)

make_ridge_plot <- function(model, plot_title = "") {

  newdata <- expand_grid(
    ons_poa = c("p", "t", "k"),
    coda_vc = c("0", "1")
  )
  newdata$ons_poa <- factor(newdata$ons_poa, levels = levels(data_mod$ons_poa))
  newdata$coda_vc <- factor(newdata$coda_vc, levels = levels(data_mod$coda_vc))

  preds <- posterior_epred(model, newdata = newdata,
                            allow_new_levels = TRUE, re_formula = NA)

  pred_summary <- newdata %>%
    mutate(
      estimate = apply(preds, 2, mean),
      lower    = apply(preds, 2, quantile, 0.025),
      upper    = apply(preds, 2, quantile, 0.975)
    )

  pred_summary$poa_coda <- factor(
    interaction(pred_summary$coda_vc, pred_summary$ons_poa),
    levels = c("0.p", "1.p", "0.t", "1.t", "0.k", "1.k")
  )

  ggplot() +
    geom_density_ridges(
      data = data_mod,
      aes(x = vot, y = poa_coda, fill = poa_coda, linetype = coda_vc_factor),
      alpha = 0.8, scale = 0.8
    ) +
    scale_linetype_manual(
      name = "Coda voicing",
      values = c("Voiceless" = "dashed", "Voiced" = "solid")
    ) +
    geom_point(
      data = pred_summary,
      aes(x = estimate, y = poa_coda, color = poa_coda),
      size = 3, position = position_nudge(y = 0.2)
    ) +
    geom_errorbarh(
      data = pred_summary,
      aes(x = estimate, y = poa_coda, xmin = lower, xmax = upper, color = poa_coda),
      height = 0.1, position = position_nudge(y = 0.2)
    ) +
    scale_fill_manual(values = c(
      "0.p" = "#B71C1C", "1.p" = "#FFCDD2",
      "0.t" = "#1B5E20", "1.t" = "#C8E6C9",
      "0.k" = "#0D47A1", "1.k" = "#BBDEFB"
    )) +
    scale_color_manual(values = c(
      "0.p" = "#67000D", "1.p" = "#A50F15",
      "0.t" = "#00441B", "1.t" = "#006D2C",
      "0.k" = "#08306B", "1.k" = "#08519C"
    )) +
    annotate("text", x = 10, y = 1.5, label = "/p/", hjust = 1.2, vjust = 0.5,
             size = 4, fontface = "bold") +
    annotate("text", x = 10, y = 3.5, label = "/t/", hjust = 1.2, vjust = 0.5,
             size = 4, fontface = "bold") +
    annotate("text", x = 10, y = 5.5, label = "/k/", hjust = 1.2, vjust = 0.5,
             size = 4, fontface = "bold") +
    labs(x = "VOT (ms)", y = "", title = plot_title) +
    xlim(0, 200) +
    coord_cartesian(ylim = c(0.5, 6.5)) +
    theme_minimal() +
    guides(fill = "none", color = "none",
           linetype = guide_legend(title = "Coda voicing", direction = "vertical")) +
    theme(
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
      legend.position = c(0.85, 0.20),
      legend.background = element_rect(fill = "white", color = "black", linewidth = 0.3),
      panel.grid.major.y = element_blank(),
      panel.grid.minor.y = element_blank(),
      plot.margin = margin(t = 15, r = 5, b = 0, l = 5, unit = "pt")
    ) +
    scale_y_discrete(expand = expansion(mult = c(0.05, 0.15)))
}
```

## Default-prior model

```{r ridge-default, fig.width=7, fig.height=5}
p_default <- make_ridge_plot(bayes_model_maximal_default, plot_title = "Default priors")
print(p_default)
ggsave("ridge_default.png", plot = p_default, bg = "white", width = 7, height = 5, dpi = 300)
```

## Informed-prior model

```{r ridge-informed, fig.width=7, fig.height=5}
p_informed <- make_ridge_plot(bayes_model_maximal_informed, plot_title = "")
print(p_informed)
ggsave("ridge_informed.png", plot = p_informed, bg = "white", width = 7, height = 5, dpi = 300)
```
