---
title: "Burst Voice: Onset VOT ~ Place of Articulation x Coda Voicing"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Data analysis

## Setup and data preparation

`ons_poa` is coded once here as a factor with levels/labels
`c("p", "t", "k")`, and `coda_vc` once as a factor with levels
`c(0, 1)`. Every plot, model, and prediction grid in the rest of this
document inherits this coding and reference level (`/p/`), rather than
each section re-deriving place-of-articulation labels independently.

```{r libraries}
library(ggplot2)
library(tidyverse)
library(dplyr)
library(ggridges)
library(lme4)
library(brms)
library(marginaleffects)
```

```{r data-prep}
data <- read.csv("/Users/chandan/Desktop/GitHub/Burst_voice/Data/data_12-11-24.csv",
                  header = TRUE)

data <- data %>%
  mutate(
    ons_poa = factor(ons_poa, levels = c(1, 2, 3), labels = c("p", "t", "k")),
    coda_vc = factor(coda_vc, levels = c(0, 1)),
    vot_ratio = vot / vdur,
    CVdur = vot + vdur
  )

# single data frame used throughout; kept as a separate name only
# for readability in code carried over from earlier drafts
data_mod <- data

number_of_subs <- length(unique(data$sub))
number_of_subs
```

## Preliminary plotting

### By onset POA

This is a first pass at looking at the data.

```{r consistency-check}
# Within-condition consistency (coefficient of variation) check
consistency_stats <- data %>%
  group_by(sub, ons_poa, coda_vc, vowel) %>%
  summarise(
    n_obs = n(),
    mean_CVdur = mean(CVdur, na.rm = TRUE),
    sd_CVdur = sd(CVdur, na.rm = TRUE),
    cv = sd_CVdur / mean_CVdur * 100,
    .groups = "drop"
  ) %>%
  filter(n_obs > 1)

consistency_stats %>%
  arrange(desc(cv)) %>%
  head(20)
```

```{r summary-table}
# Summary table of mean VOT by place and coda voicing
summary_table <- data %>%
  group_by(ons_poa, coda_vc) %>%
  summarize(
    mean_vot = mean(vot, na.rm = TRUE),
    vot_range = paste0(min(vot, na.rm = TRUE), " to ", max(vot, na.rm = TRUE))
  ) %>%
  ungroup()

print(summary_table)
```

```{r plot-overall, fig.width=8, fig.height=5}
# Boxplot: VOT by onset POA and coda voicing
plot_overall <- ggplot(data, aes(x = ons_poa, y = vot, fill = coda_vc)) +
  geom_boxplot(notch = TRUE, outlier.shape = NA) +
  theme(axis.text.x = element_blank()) +
  labs(x = NULL, y = "CV duration (s)") +
  facet_wrap(~ons_poa, scales = "free") +
  scale_y_continuous(limits = c(0.015, 0.18)) +
  scale_fill_discrete(name = "Coda voice", labels = c("-vc", "+vc"))

print(plot_overall)
```

```{r plot-overall-by-sub, fig.width=10, fig.height=8}
plot_overall + facet_wrap(~sub)
```

```{r plot-cvdur, fig.width=8, fig.height=5}
# Boxplot: CV duration by onset POA and coda voicing
plot_overall_CVdur <- ggplot(data, aes(x = ons_poa, y = CVdur, fill = coda_vc)) +
  geom_boxplot(notch = TRUE, outlier.shape = NA) +
  theme(axis.text.x = element_blank()) +
  labs(x = NULL, y = "CV duration (s)") +
  facet_wrap(~ons_poa, scales = "free") +
  scale_fill_discrete(name = "Coda voice", labels = c("-vc", "+vc"))

print(plot_overall_CVdur)
```

```{r plot-vot-poa, fig.width=10, fig.height=8}
# Boxplot: VOT by onset POA, faceted by subject
plot_vot_poa <- ggplot(data, aes(x = vot, y = ons_poa, fill = coda_vc)) +
  geom_boxplot(notch = TRUE, outlier.shape = NA, width = 0.5) +
  theme_minimal() +
  labs(x = "VOT(s)", y = "Onset POA", fill = "Coda voice") +
  scale_fill_discrete(labels = c("[-vc]", "[+vc]"))

print(plot_vot_poa + facet_wrap(~sub))
```

```{r plot-vdur-poa, fig.width=8, fig.height=5}
# Boxplot: vowel duration by onset POA, faceted by subject
plot_vdur_poa <- ggplot(data, aes(x = vdur, y = ons_poa, fill = coda_vc)) +
  geom_boxplot(notch = TRUE, outlier.shape = NA, width = 0.5) +
  theme_minimal() +
  labs(x = "Vdur(s)", y = "Onset POA", fill = "Coda voice") +
  scale_x_continuous(limits = c(0, 0.3), expand = c(0, 0)) +
  scale_fill_discrete(labels = c("[-vc]", "[+vc]"))

plot_vdur_poa
```

```{r plot-vdur-poa-by-sub, fig.width=10, fig.height=8}
plot_vdur_poa + facet_wrap(~sub)
```

### By onset POA and vowel

Ridge plots (stacked density plots) for voiceless and voiced coda
tokens by POA and vowel.

```{r plot-vowel-boxplot, fig.width=9, fig.height=6}
plot_vowel <- ggplot(data, aes(x = vowel, y = vot, fill = coda_vc)) +
  geom_boxplot(notch = TRUE) +
  labs(y = "VOT(s)", x = "Vowel", fill = "Coda voicing") +
  facet_wrap(~ons_poa, scale = "free")
plot_vowel
```

```{r ridge-labels}
# ons_poa is already factor-coded p/t/k, so these labellers map from
# that coding to display symbols only
poa_labels <- c("p" = "[p\u02b0]", "t" = "[t\u02b0]", "k" = "[k\u02b0]")
vowel_labels <- c("ae" = "\u00e6", "aw" = "\u0254", "eh" = "\u025b",
                   "ih" = "\u026a", "uh" = "\u028c")
```

```{r ridge-by-vowel, fig.width=12, fig.height=6}
plot_vowel_ridges <- ggplot(data, aes(x = vot, y = coda_vc, fill = coda_vc)) +
  geom_density_ridges(alpha = 0.7) +
  scale_x_continuous(breaks = c(0, 0.1, 0.2), limits = c(0, NA)) +
  scale_y_discrete(labels = c("0" = "Voiceless", "1" = "Voiced")) +
  scale_fill_manual(
    name = "Coda",
    values = c("0" = "#F8766D", "1" = "#00BFC4"),
    labels = c("0" = "Voiceless", "1" = "Voiced")
  ) +
  labs(x = "VOT(s)", y = "") +
  facet_grid(vowel ~ ons_poa, scales = "free_x",
             labeller = labeller(ons_poa = poa_labels, vowel = vowel_labels)) +
  theme_ridges() +
  theme(axis.text.y = element_blank(),
        axis.title.x = element_text(hjust = 0.5),
        strip.text = element_text(margin = margin(t = 4, b = 4)),
        strip.text.x = element_text(margin = margin(t = 3, b = 4)),
        strip.text.y = element_text(margin = margin(l = 3, r = 3)),
        panel.spacing = unit(0.3, "cm"))

print(plot_vowel_ridges)
ggsave("plot_ridges.png", plot = plot_vowel_ridges, width = 12, height = 6, dpi = 300)
```

```{r ridge-collapsed, fig.width=9, fig.height=6}
# Same ridge plot, collapsed across vowel
plot_ridges <- ggplot(data, aes(x = vot, y = coda_vc, fill = coda_vc)) +
  geom_density_ridges(alpha = 0.7) +
  scale_x_continuous(breaks = c(0, 0.05, 0.1, 0.15, 0.2), limits = c(0, NA)) +
  scale_y_discrete(labels = c("0" = "Voiceless", "1" = "Voiced"), expand = c(0, 0)) +
  scale_fill_manual(
    name = "Coda",
    values = c("0" = "#F8766D", "1" = "#00BFC4"),
    labels = c("0" = "Voiceless", "1" = "Voiced")
  ) +
  labs(x = "VOT(s)", y = "") +
  facet_wrap(~ons_poa, scales = "free_x",
             labeller = labeller(ons_poa = poa_labels)) +
  theme_ridges() +
  theme(axis.text.y = element_blank(),
        axis.title.x = element_text(hjust = 0.5),
        strip.text = element_text(margin = margin(t = 4, b = 4)),
        strip.text.x = element_text(margin = margin(t = 3, b = 4)),
        panel.spacing = unit(0.3, "cm"),
        legend.position = c(0.85, 0.85),
        legend.background = element_rect(fill = "white", color = "black", linewidth = 0.3),
        legend.margin = margin(t = 5, r = 5, b = 5, l = 5),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 10))

plot_ridges
```

```{r density-by-poa-vowel, fig.width=12, fig.height=9}
# Overlapping density plots by POA and vowel
plot_vowel_density <- ggplot(data, aes(x = vot, fill = coda_vc)) +
  geom_density(alpha = 0.6) +
  labs(x = "VOT(s)", y = "Density", fill = "Coda voicing") +
  facet_wrap(~ons_poa + vowel, scales = "free")
plot_vowel_density
```

It looks like there is a clear effect of coda voicing on voiceless
onset VOT.

## Basic models

```{r lmer-full}
model_full <- lmer(vot ~ ons_poa * coda_vc * vowel + (1 | sub), data = data_mod)
summary(model_full)
```

Removing the vowel term (reference level `/p/`, matching the
`ons_poa` factor coding set above -- no re-recoding needed here):

```{r lmer-reduced}
model <- lmer(vot ~ ons_poa * coda_vc + (1 | sub), data = data_mod)
summary(model)
```

### Plot of the model

```{r plot-model, fig.width=9, fig.height=5}
p_lmer <- ggplot(data_mod, aes(x = vdur, y = vot, color = coda_vc)) +
  geom_point() +
  geom_smooth(method = lm) +
  xlab("Vowel dur (ms)") + ylab("VOT (ms)")
p_lmer + facet_wrap(~ons_poa)
```

### GLMM: predicting coda voicing from onset VOT and POA

```{r glmm}
model_coda <- glmer(
  coda_vc ~ vot * ons_poa + (1 | sub),
  data = data,
  family = binomial(link = "logit")
)

summary(model_coda)
```

```{r glmm-predictions, fig.width=8, fig.height=5}
# Predictions across a range of VOT values, for each POA
vot_range <- seq(from = 0, to = 0.25, by = 0.005)
sub_ref <- unique(data$sub)[1]

plot_data <- expand_grid(
  vot = vot_range,
  ons_poa = levels(data$ons_poa),   # already "p", "t", "k", in that order
  sub = sub_ref
)
plot_data$ons_poa <- factor(plot_data$ons_poa, levels = levels(data$ons_poa))
plot_data$coda_vc <- NA

plot_data$predicted_prob <- predict(model_coda, newdata = plot_data,
                                     type = "response", re.form = ~(1 | sub))

vc_pred <- ggplot(plot_data, aes(x = vot, y = predicted_prob, color = ons_poa)) +
  geom_line(linewidth = 2) +
  labs(
    title = "",
    x = "VOT (s)",
    y = "Predicted Probability of Voiced Coda",
    color = "Onset POA"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.title = element_text(size = 14),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )

vc_pred
ggsave("vc_pred.png", plot = vc_pred, bg = "white", width = 7, height = 4, dpi = 300)
```

# Bayesian mixed-effects model of VOT ~ coda voicing

Reference level for `ons_poa` is `/p/`, set once in the data
preparation section above.

```{r bayes-prior-check}
# Confirm reference level / contrast coding before fitting
contrasts(data_mod$ons_poa)
get_prior(vot ~ ons_poa * coda_vc + (1 | sub), data = data_mod, family = gaussian())
```

## Default (weakly informative) prior model

```{r bayes-default, results='hide'}
bayes_model_default <- brm(
  vot ~ ons_poa * coda_vc + (1 | sub),
  data   = data_mod,
  family = gaussian(),
  chains = 4,
  iter   = 2000,
  cores  = 4,
  seed   = 123
)
```

```{r bayes-default-summary, fig.width=9, fig.height=7}
# digits = 4 gives sub-millisecond precision (values are in seconds,
# so 4 decimal places = 0.1 ms resolution), rather than the default
# 2 decimal places (10 ms resolution)
print(summary(bayes_model_default), digits = 4)
plot(bayes_model_default)
```

```{r bayes-default-fixef}
# Programmatic access to full-precision estimates, useful for
# generating tables directly rather than transcribing console output
fixef(bayes_model_default, digits = 4)
VarCorr(bayes_model_default)
```

## Informed prior model

Priors are split by epistemic status: place-of-articulation effects
are strongly informed by the established cross-linguistic VOT
literature, while the coda-voicing effect (the focal, untested
hypothesis) is given a weakly informative, zero-centered prior rather
than a value read off this dataset's own `lmer` estimate (~12ms), to
avoid double-dipping / circularity.

```{r bayes-informed-priors}
informed_priors <- c(
  # Strongly informed: established literature on POA and VOT
  prior(normal(30, 10), class = Intercept),              # /p/ baseline
  prior(normal(15, 8),  class = b, coef = ons_poat),      # /t/ effect
  prior(normal(30, 12), class = b, coef = ons_poak),      # /k/ effect

  # Weakly informed / regularizing: the focal, untested hypothesis
  prior(normal(0, 20), class = b, coef = "coda_vc1"),
  prior(normal(0, 15), class = b, coef = "ons_poat:coda_vc1"),
  prior(normal(0, 15), class = b, coef = "ons_poak:coda_vc1"),

  # Standard weakly informative priors for variance components
  prior(normal(0, 20), class = sd),
  prior(normal(0, 25), class = sigma)
)
```

```{r bayes-informed-fit, results='hide'}
bayes_model_informed <- brm(
  vot ~ ons_poa * coda_vc + (1 | sub),
  data   = data_mod,
  family = gaussian(),
  prior  = informed_priors,
  chains = 4,
  iter   = 2000,
  cores  = 4,
  seed   = 123
)
```

```{r bayes-informed-summary}
prior_summary(bayes_model_informed)
print(summary(bayes_model_informed), digits = 4)
```

```{r bayes-informed-fixef}
fixef(bayes_model_informed, digits = 4)
VarCorr(bayes_model_informed)
```

```{r bayes-informed-ppcheck, fig.width=7, fig.height=5}
pp_check(bayes_model_informed, type = "hist", ndraws = 100) +
  ggtitle("Prior Predictive Check: Do simulated VOTs look reasonable?")
```

## Model comparison

```{r bayes-loo}
loo_default  <- loo(bayes_model_default)
loo_informed <- loo(bayes_model_informed)
loo_compare(loo_default, loo_informed)
```

Direct posterior comparison for the focal coefficient, now directly
comparable since both models share the `/p/` reference level:

```{r bayes-posterior-compare}
posterior_summary(bayes_model_default,  variable = "b_coda_vc1")
posterior_summary(bayes_model_informed, variable = "b_coda_vc1")
```

### Full-precision estimates for all fixed effects

`posterior_summary()` returns unrounded values for every parameter,
not just `coda_vc1`. Extracting all of them at once, in milliseconds,
gives a single unrounded source to draw on when building tables (e.g.,
for the manuscript), rather than reading rounded values off the
default `summary()` printout.

```{r bayes-full-precision-ms}
# Fixed effects, converted to ms, full precision, both models side by side
default_fx  <- as.data.frame(posterior_summary(bayes_model_default))  %>%
  rownames_to_column("parameter") %>%
  filter(str_starts(parameter, "b_")) %>%
  mutate(across(where(is.numeric), ~ .x * 1000)) %>%
  rename_with(~ paste0("default_", .x), -parameter)

informed_fx <- as.data.frame(posterior_summary(bayes_model_informed)) %>%
  rownames_to_column("parameter") %>%
  filter(str_starts(parameter, "b_")) %>%
  mutate(across(where(is.numeric), ~ .x * 1000)) %>%
  rename_with(~ paste0("informed_", .x), -parameter)

fixef_comparison <- left_join(default_fx, informed_fx, by = "parameter") %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))

fixef_comparison
```

# Ridge plots with posterior predictions

Ridge plots with posterior predictions overlaid, for both Bayesian
models, on the shared `/p/` reference level.

```{r ridge-pred-setup}
data_mod$poa_coda <- factor(
  interaction(data_mod$coda_vc, data_mod$ons_poa),
  levels = c("0.p", "1.p", "0.t", "1.t", "0.k", "1.k")
)

data_mod$coda_vc_factor <- factor(
  data_mod$coda_vc,
  levels = c("1", "0"),
  labels = c("Voiced", "Voiceless")
)

make_ridge_plot <- function(model, plot_title = "") {

  newdata <- expand_grid(
    ons_poa = c("p", "t", "k"),
    coda_vc = c("0", "1")
  )
  newdata$ons_poa <- factor(newdata$ons_poa, levels = levels(data_mod$ons_poa))
  newdata$coda_vc <- factor(newdata$coda_vc, levels = levels(data_mod$coda_vc))

  preds <- posterior_epred(model, newdata = newdata,
                            allow_new_levels = TRUE, re_formula = NA)

  pred_summary <- newdata %>%
    mutate(
      estimate = apply(preds, 2, mean),
      lower    = apply(preds, 2, quantile, 0.025),
      upper    = apply(preds, 2, quantile, 0.975)
    )

  pred_summary$poa_coda <- factor(
    interaction(pred_summary$coda_vc, pred_summary$ons_poa),
    levels = c("0.p", "1.p", "0.t", "1.t", "0.k", "1.k")
  )

  ggplot() +
    geom_density_ridges(
      data = data_mod,
      aes(x = vot, y = poa_coda, fill = poa_coda, linetype = coda_vc_factor),
      alpha = 0.8, scale = 0.8
    ) +
    scale_linetype_manual(
      name = "Coda voicing",
      values = c("Voiceless" = "dashed", "Voiced" = "solid")
    ) +
    geom_point(
      data = pred_summary,
      aes(x = estimate, y = poa_coda, color = poa_coda),
      size = 3, position = position_nudge(y = 0.2)
    ) +
    geom_errorbarh(
      data = pred_summary,
      aes(x = estimate, y = poa_coda, xmin = lower, xmax = upper, color = poa_coda),
      height = 0.1, position = position_nudge(y = 0.2)
    ) +
    scale_fill_manual(values = c(
      "0.p" = "#B71C1C", "1.p" = "#FFCDD2",
      "0.t" = "#1B5E20", "1.t" = "#C8E6C9",
      "0.k" = "#0D47A1", "1.k" = "#BBDEFB"
    )) +
    scale_color_manual(values = c(
      "0.p" = "#67000D", "1.p" = "#A50F15",
      "0.t" = "#00441B", "1.t" = "#006D2C",
      "0.k" = "#08306B", "1.k" = "#08519C"
    )) +
    annotate("text", x = 0.01, y = 1.5, label = "p\u02b0", hjust = 1.2, vjust = 0.5,
             size = 4, fontface = "bold") +
    annotate("text", x = 0.01, y = 3.5, label = "t\u02b0", hjust = 1.2, vjust = 0.5,
             size = 4, fontface = "bold") +
    annotate("text", x = 0.01, y = 5.5, label = "k\u02b0", hjust = 1.2, vjust = 0.5,
             size = 4, fontface = "bold") +
    labs(x = "VOT (s)", y = "", title = plot_title) +
    xlim(0, 0.2) +
    coord_cartesian(ylim = c(0.5, 6.5)) +
    theme_minimal() +
    guides(fill = "none", color = "none",
           linetype = guide_legend(title = "Coda voicing", direction = "vertical")) +
    theme(
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
      legend.position = c(0.85, 0.20),
      legend.background = element_rect(fill = "white", color = "black", linewidth = 0.3),
      panel.grid.major.y = element_blank(),
      panel.grid.minor.y = element_blank(),
      plot.margin = margin(t = 15, r = 5, b = 0, l = 5, unit = "pt")
    ) +
    scale_y_discrete(expand = expansion(mult = c(0.05, 0.15)))
}
```

## Default-prior model

```{r ridge-default, fig.width=7, fig.height=5}
p_default <- make_ridge_plot(bayes_model_default, plot_title = "Default priors")
print(p_default)
ggsave("ridge_default.png", plot = p_default, bg = "white", width = 7, height = 5, dpi = 300)
```

## Informed-prior model

```{r ridge-informed, fig.width=7, fig.height=5}
p_informed <- make_ridge_plot(bayes_model_informed, plot_title = "Informed priors")
print(p_informed)
ggsave("ridge_informed.png", plot = p_informed, bg = "white", width = 7, height = 5, dpi = 300)
```
