---
title: "Data analysis"
output: 
  html_document:
    includes:
      in_header: "favicon.html"
    theme: paper
    toc: true
    toc_float: true
    collapsed: false
    number_sections: false
    toc_depth: 3
---

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>

<style type="text/css">
.title {
  display: none;
}

#getting-started img {
  margin-right: 10px;
}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Preliminary plotting

# By Onset POA

This is just a first pass at looking at the data so far.

```{r, warning=FALSE}
data <- read.csv("/Users/chandannarayan/GitHub/Burst_voice/Data/data_12-11-24.csv", header=TRUE)
data$coda_vc <- as.character(data$coda_vc) #change class of coda_vc
data$ons_poa <- as.character(data$ons_poa) #change class of ons_poa
data$vot_ratio <- (data$vot)/(data$vdur)

number_of_subs <- length(unique(data$sub))
number_of_subs

library(ggplot2)
library(tidyverse)

# Create a summary table with the mean and range of 'vot' for each 'ons_poa'
summary_table <- data %>%
  group_by(ons_poa, coda_vc) %>%
  summarize(
    mean_vot = mean(vot, na.rm = TRUE),
    vot_range = paste0(min(vot, na.rm = TRUE), " to ", max(vot, na.rm = TRUE))
  ) %>%
  ungroup()  # Ungroup after summarizing

# Print the summary table
print(summary_table)

plot_overall <- ggplot(data, aes(x = ons_poa, y = vot, fill = coda_vc)) +
  geom_boxplot(notch = TRUE, outlier.shape = NA) +
  # Remove x-axis tick labels
  theme(axis.text.x = element_blank()) +
  # Remove x-axis label and update y-axis label
  labs(x = NULL, y = "VOT(s)") +
  # Change facet labels
  facet_wrap(~ons_poa, scales = "free", labeller = labeller(ons_poa = c("1" = "p", "2" = "t", "3" = "k"))) +
  # Set consistent y-axis range
  scale_y_continuous(limits = c(0.015, 0.18)) +
  # Update the legend
  scale_fill_discrete(
    name = "Coda voice",
    labels = c("-vc", "+vc")
  ) 
# + stat_summary(fun = median,geom = "text",aes(label = paste("Median:", round(..y.., 3))),  # Label with median values
#    vjust = -0.5,  # Vertical position of the text (adjust as needed)
#    position = position_dodge(width = 0.75),  # Dodge the text for better visibility
#    show.legend = FALSE ) # Do not show in legend
  
plot_overall

# Plot by sub
plot_vot_poa <- ggplot(data, aes(x = vot, y = as.factor(ons_poa), fill = as.factor(coda_vc))) +
  geom_boxplot(notch = TRUE, outlier.shape = NA, width = 0.5) +  # Notched boxplots
  theme_minimal() +  # Clean aesthetic
  labs(x = "VOT(s)", y = "Onset POA", fill = "Coda voice") +
  scale_y_discrete(labels = c("1" = "p", "2" = "t", "3" = "k")) +  # Custom y-axis labels
  #scale_x_continuous(limits = c(0, 2), expand = c(0, 0)) +  # Adjust x-axis range
  scale_fill_discrete(
    labels = c("0" = "[-vc]", "1" = "[+vc]")  # Custom legend labels
  )

print(plot_vot_poa + facet_wrap(~sub))

plot_vdur_poa <- ggplot(data, aes(x = vdur, y = as.factor(ons_poa), fill = as.factor(coda_vc))) +
  geom_boxplot(notch = TRUE, outlier.shape = NA, width = 0.5) +  # Notched boxplots
  theme_minimal() +  # Clean aesthetic
  labs(x = "Vdur(s)", y = "Onset POA", fill = "Coda voice") +
  scale_y_discrete(labels = c("1" = "p", "2" = "t", "3" = "k")) +  # Custom y-axis labels
  scale_x_continuous(limits = c(0, 0.3), expand = c(0, 0)) +  # Adjust x-axis range
  scale_fill_discrete(
    labels = c("0" = "[-vc]", "1" = "[+vc]")  # Custom legend labels
  )
plot_vdur_poa
plot_vdur_poa + facet_wrap(~sub)

plot_overall + facet_wrap(~sub)


```

# By Onset POA and Vowel

I'm using a "ridge plot" here which is a stacked density plot for voiceless and voiced coda tokens by POA and vowel. 

```{r, warning=FALSE}
plot_vowel <- ggplot(data, aes(x = vowel,y = vot, fill = coda_vc)) + geom_boxplot(notch=TRUE) + 
  labs(y="VOT(s)", x="Vowel", fill="Coda voicing")+
  facet_wrap(~ons_poa, scale = "free")
plot_vowel

## Ridge plot with medians
library(dplyr)
library(ggridges)

# New facet label names for POA variable
poa.labs <- c("p", "t", "k")
names(poa.labs) <- c("1", "2", "3")

# Create custom labels
poa_labels <- c("1" = "[pʰ]",
                "2" = "[tʰ]",
                "3" = "[kʰ]")

vowel_labels <- c("ae" = "æ",
                  "aw" = "ɔ",
                  "eh" = "ɛ",
                  "ih" = "ɪ",
                  "uh" = "ʌ")

plot_vowel_ridges <- ggplot(data, aes(x = vot, y = coda_vc, fill = coda_vc)) + 
  geom_density_ridges(alpha = 0.7) + 
  #geom_vline(data = medians_data, aes(xintercept = median_vot), 
             #linetype = "dotted", color = "black", linewidth = 0.5) +
  scale_x_continuous(breaks = c(0, 0.1, 0.2), limits = c(0, NA)) +
  scale_y_discrete(labels = c("0" = "Voiceless", "1" = "Voiced")) +
  scale_fill_manual(
    name = "Coda", 
    values = c("0" = "#F8766D", "1" = "#00BFC4"),
    labels = c("0" = "Voiceless", "1" = "Voiced")
  ) +
  labs(x = "VOT(s)", y = "") +
  facet_grid(vowel ~ ons_poa, scales = "free_x",
             labeller = labeller(ons_poa = poa_labels, vowel = vowel_labels)) +
  theme_ridges() + 
  theme(axis.text.y = element_blank(),
        axis.title.x = element_text(hjust = 0.5),
        strip.text = element_text(margin = margin(t = 4, b = 4)),
        strip.text.x = element_text(margin = margin(t = 3, b = 4)),  
        strip.text.y = element_text(margin = margin(l = 3, r = 3)),
        panel.spacing = unit(0.3, "cm"))
print(plot_vowel_ridges)

## Remove the vowel facet:

# Calculate means without vowel grouping
means_data <- data %>%
  group_by(ons_poa, coda_vc) %>%
  summarise(mean_vot = mean(vot, na.rm = TRUE), .groups = 'drop')
# I'm not using the means right now

plot_ridges <- ggplot(data, aes(x = vot, y = coda_vc, fill = coda_vc)) + 
  geom_density_ridges(alpha = 0.7) + 
  scale_x_continuous(breaks = c(0, 0.05, 0.1, 0.15, 0.2), limits = c(0, NA)) +
  scale_y_discrete(labels = c("0" = "Voiceless", "1" = "Voiced"), expand = c(0,0)) +
  scale_fill_manual(
    name = "Coda", 
    values = c("0" = "#F8766D", "1" = "#00BFC4"),
    labels = c("0" = "Voiceless", "1" = "Voiced")
  ) +
  labs(x = "VOT(s)", y = "") +
  facet_wrap(~ ons_poa, scales = "free_x",
             labeller = labeller(ons_poa = poa_labels)) +
  theme_ridges() + 
  theme(axis.text.y = element_blank(),
        axis.title.x = element_text(hjust = 0.5),
        strip.text = element_text(margin = margin(t = 4, b = 4)),
        strip.text.x = element_text(margin = margin(t = 3, b = 4)),  
        panel.spacing = unit(0.3, "cm"),
        legend.position = c(0.85, 0.85),  # Position legend inside plot area
        legend.background = element_rect(fill = "white", color = "black", size = 0.3),
        legend.margin = margin(t = 5, r = 5, b = 5, l = 5),
        legend.title = element_text(size=12),
        legend.text = element_text(size=10))

# To widen the image when you save it, use:
ggsave("plot_ridges.png", plot = plot_vowel_ridges, 
       width = 12, height = 6, dpi = 300)
plot_ridges

## overlapping density plots

plot_vowel_density <- ggplot(data, aes(x = vot, fill = coda_vc)) + 
  geom_density(alpha = 0.6) + 
  labs(x = "VOT(s)", y = "Density", fill = "Coda voicing") +
  facet_wrap(~ons_poa + vowel, scales = "free")
plot_vowel_density
```

It looks like there is a clear effect of coda voicing on voiceless onset VOT.

# Basic models

```{r, warning=FALSE}
library(lme4)
library(tidyverse)
data_mod <- data %>%
  #rename POA labels from 1,2,3 to p,t,k
  mutate(ons_poa = recode(ons_poa, "1" = "p", "2" = "t", "3" = "k"))

model_full <- lmer(vot ~ ons_poa * coda_vc * vowel + (1 | sub), data = data_mod)
summary(model_full)
```
Remove the vowel term:
```{r}
data_mod <- data %>%
  mutate(ons_poa = recode(ons_poa, "1" = "p-", "2" = "t-", "3" = "k-"))

model <- lmer(vot ~ ons_poa * coda_vc + (1 | sub), data = data_mod)
summary(model)

```
# Plot of the model
```{r, warning=FALSE}
p <- ggplot(data_mod, aes(x=vdur, y=vot, color=coda_vc)) +
  geom_point(outlier.shape = NA) +
  geom_smooth(method=lm) + xlab("Vowel dur (ms)") + ylab("VOT (ms)")
p + facet_wrap(~ons_poa)
```

```{r, warning=FALSE}
# Load required package
library(lme4)

# Convert coda_vc to a factor if it's not already
data$coda_vc <- factor(data$coda_vc, levels = c(0,1))

# Fit the GLMM
model_coda <- glmer(
  coda_vc ~ vot * ons_poa + (1 | sub), 
  data = data, 
  family = binomial(link = "logit")
)

# Summary of the model
summary(model_coda)

# Generate new data for predictions
vot_range <- seq(from = 0, to = 0.25, by = 0.005)
ons_poa_levels <- c("1", "2", "3")
#vowel_levels <- c("ae","uh","aw","eh","ih")
#sub_levels <- c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19)
sub = unique(data$sub)[1]

# Create a data frame for plotting
plot_data <- expand.grid(
  vot = vot_range,
  ons_poa = ons_poa_levels,
  #vowel = vowel_levels,
  sub = sub
)

# Add a dummy coda_vc variable (not used in prediction but avoids the error)
plot_data$coda_vc <- NA

# Add predictions from the model
plot_data$predicted_prob <- predict(model_coda, newdata = plot_data, type = "response", re.form = ~(1 | sub))



vc_pred <- ggplot(plot_data, aes(x = vot, y = predicted_prob, color = ons_poa )) +
  geom_line(size = 2) +
  #geom_jitter(data = data, aes(x = vot, y = as.numeric(coda_vc), color = ons_poa), 
   #           width = 0, height = 0.05, alpha = 0.5, shape = 16) +  # Jitter actual points
scale_color_discrete(
    labels = c("1" = "p", "2" = "t", "3" = "k")  # Custom legend labels
  ) +
  scale_fill_discrete(
    labels = c("1" = "p", "2" = "t", "3" = "k")  # Custom legend labels for fill
  ) +
  labs(
    title = "",
    x = "VOT (s)",
    y = "Predicted Probability of Voiced Coda",
    color = "Onset POA"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.title = element_text(size = 14),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )

vc_pred
ggsave("vc_pred.png", plot = vc_pred, bg = 'white', width = 7, height =4, dpi=300)

```

# Bayesian mixed effects model of VOT ~ coda voicing

```{r}

library(brms)
library(tidyverse)

data_mod <- data %>%
  #rename POA labels from 1,2,3 to p,t,k
  mutate(ons_poa = recode(ons_poa, "1" = "p", "2" = "t", "3" = "k"))

# Bayesian version of your model
bayes_model <- brm(vot ~ ons_poa * coda_vc + (1 | sub),
                   data = data_mod,
                   family = gaussian(),
                   chains = 4,
                   iter = 2000,
                   cores = 4)

# Check model
summary(bayes_model)
plot(bayes_model)

library(marginaleffects)
library(ggplot2)
library(ggridges)

# Get predictions
newdata <- expand_grid(
  ons_poa = c("k", "p", "t"),
  coda_vc = c(0, 1)
)

preds <- posterior_epred(bayes_model, newdata = newdata,
                         allow_new_levels = TRUE, re_formula = NA)

pred_summary <- newdata %>%
  mutate(
    estimate = apply(preds, 2, mean),
    lower = apply(preds, 2, quantile, 0.025),
    upper = apply(preds, 2, quantile, 0.975)
  )

# Reorder the data and create color scheme
data_mod$poa_coda <- factor(interaction(data_mod$coda_vc, data_mod$ons_poa),
                           levels = c("0.p", "1.p", "0.t", "1.t", "0.k", "1.k"))
data_mod$coda_vc_factor <- factor(data_mod$coda_vc, levels = c("1", "0"), 
                                 labels = c("Voiced", "Voiceless"))
pred_summary$poa_coda <- factor(interaction(pred_summary$coda_vc, pred_summary$ons_poa),
                               levels = c("0.p", "1.p", "0.t", "1.t", "0.k", "1.k"))

# Create the plot
p <- ggplot() +
  # Ridged density plot of actual data with increased contrast
  geom_density_ridges(data = data_mod,
                      aes(x = vot, y = poa_coda, fill = poa_coda, 
                          linetype = coda_vc_factor),
                     alpha = 0.8, scale = 0.8) +
  scale_linetype_manual(name = "Coda voicing",
                       values = c("Voiceless" = "dashed", "Voiced" = "solid")) +
  # Overlay predictions with darker colors
  geom_point(data = pred_summary,
            aes(x = estimate, y = poa_coda, color = poa_coda),
            size = 3, position = position_nudge(y = 0.2)) +
  geom_errorbarh(data = pred_summary,
                aes(x = estimate, y = poa_coda,
                    xmin = lower, xmax = upper, color = poa_coda),
                height = 0.1, position = position_nudge(y = 0.2)) +
  # Custom colors with higher contrast between voiced/voiceless
  scale_fill_manual(values = c("0.p" = "#B71C1C", "1.p" = "#FFCDD2",
                              "0.t" = "#1B5E20", "1.t" = "#C8E6C9",
                              "0.k" = "#0D47A1", "1.k" = "#BBDEFB")) +
  # Much darker colors for predictions
  scale_color_manual(values = c("0.p" = "#67000D", "1.p" = "#A50F15",
                               "0.t" = "#00441B", "1.t" = "#006D2C",
                               "0.k" = "#08306B", "1.k" = "#08519C")) +
  # Add POA labels on the left
  annotate("text", x = 0.01, y = 1.5, label = "pʰ", hjust = 1.2, vjust = 0.5, size = 4, fontface = "bold") +
  annotate("text", x = 0.01, y = 3.5, label = "tʰ", hjust = 1.2, vjust = 0.5, size = 4, fontface = "bold") +
  annotate("text", x = 0.01, y = 5.5, label = "kʰ", hjust = 1.2, vjust = 0.5, size = 4, fontface = "bold") +
  # Formatting with restricted x-axis
  labs(x = "VOT (s)", y = "") +
  xlim(0, 0.2) +  # Restrict x-axis to 0-0.2
  coord_cartesian(ylim = c(0.5, 6.5)) +  # Extend y-axis limits
  theme_minimal() +
  guides(fill = "none", color = "none",
         linetype = guide_legend(title = "Coda voicing", direction = "vertical")) +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = c(0.85, 0.20),
        legend.background = element_rect(fill = "white", color = "black", size = 0.3),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        plot.margin = margin(t = 15, r = 5, b = 0, l = 5, unit = "pt")) +
  scale_y_discrete(expand = expansion(mult = c(0.05, 0.15)))

print(p)
```

## **Results interpretation**

#### Model Diagnostics

Rhat values: All are ≤ 1.02, which indicates excellent convergence (want < 1.1)
ESS values: All are > 200, showing adequate sampling (though some are on the lower side)

#### Random Effects (Subject Variation)
sd(Intercept): 0.03 [0.02, 0.04]

Between-subject variability in baseline VOT is 0.03 seconds
This is relatively small, suggesting subjects are fairly consistent in their baseline VOT

#### Fixed Effects (Your Main Results)
Baseline (Intercept): 0.09s [0.08, 0.10]

Mean VOT when ons_poa is at reference level and coda_vc = 0

#### Main Effects:

ons_poapM: -0.01s [-0.02, -0.01] - Clear negative effect
ons_poatM: -0.00s [-0.01, -0.00] - Very small negative effect
coda_vc1: 0.01s [0.00, 0.01] - Small positive effect

#### Interactions:

ons_poapM:coda_vc1: 0.00s [0.00, 0.01] - Minimal interaction
ons_poatM:coda_vc1: -0.00s [-0.00, 0.00] - Essentially zero

#### Residual Variation
sigma: 0.02s [0.02, 0.02]

Within-subject measurement error is 0.02 seconds

### Interpretation
Your place of articulation (ons_poa) appears to have clear effects on VOT, with both pM and tM showing shorter VOT than your reference category. The voicing of the coda (coda_vc) has a smaller positive effect. The interactions are minimal, suggesting the coda effect is consistent across places of articulation.

Perfect! Now your results make much more phonetic sense:

## **Main Findings**

**Reference Level**: `k` (velar) with baseline VOT of **0.09s**

**Place of Articulation Effects:**
- **p (labial)**: -0.01s [-0.02, -0.01] → **0.08s VOT**
- **t (coronal)**: -0.00s [-0.01, -0.00] → **~0.09s VOT** 
- **k (velar)**: 0.09s (reference)

This shows the expected **phonetic hierarchy**: **k > t > p**

## **Phonetic Interpretation**

Your results align perfectly with cross-linguistic patterns:
- **Velar stops** (k) have the **longest VOT** (0.09s)
- **Coronal stops** (t) have **intermediate VOT** (~0.09s, very close to velars)
- **Labial stops** (p) have the **shortest VOT** (0.08s)

This reflects the **aerodynamic and articulatory** differences:
- Velars have larger oral cavity volume behind constriction → longer VOT
- Labials have smaller cavity → shorter VOT
- Coronals fall between the two

## **Coda Effect**
**coda_vc1**: +0.01s [0.00, 0.01]
- Voiced codas slightly **increase** VOT of the onset stop
- This could reflect **compensatory timing** or **coarticulatory effects**

## **Statistical Strength**
The **credible intervals** for p don't include zero, showing this is a **reliable effect**. The t effect is very small and the CI barely excludes zero.

## Possible write-up

Here's how I'd write up these results for a phonetics/acoustics audience:

## **Results**

We fitted a Bayesian linear mixed-effects model to examine the effects of place of articulation and coda voicing on VOT, with random intercepts for participants. The model showed excellent convergence (all Rhat ≤ 1.02) and adequate sampling efficiency.

**Place of articulation effects.** VOT varied systematically by place of articulation, following the expected aerodynamic hierarchy. Velar stops served as the reference category with a mean VOT of 90 ms (95% CrI: [80, 100]). Labial stops showed significantly shorter VOT, with a mean decrease of 10 ms (95% CrI: [-20, -10]) relative to velars, resulting in a mean VOT of 80 ms. Coronal stops showed minimal difference from velars, with a negligible decrease of <1 ms (95% CrI: [-10, 0]), yielding a mean VOT of ~90 ms. This pattern aligns with cross-linguistic tendencies where velars exhibit the longest VOT due to their larger oral cavity volume behind the constriction, while labials show the shortest VOT.

**Coda voicing effects.** Voiced codas produced a small but reliable increase in onset VOT of 10 ms (95% CrI: [0, 10]) compared to voiceless codas. This suggests potential compensatory timing effects or coarticulatory influence of the coda on onset timing.

**Individual variation.** Between-speaker variation in baseline VOT was modest (SD = 30 ms, 95% CrI: [20, 40]), indicating relatively consistent production patterns across speakers. Within-speaker measurement error was 20 ms (95% CrI: [20, 20]).

**Interactions.** The interaction between place of articulation and coda voicing was minimal, suggesting that coda effects on VOT are consistent across places of articulation.

---
